---
title: "Advent of Code 2020"
author: "Barry Bullas"
output:
  html_document:
    toc: true
    toc_float: true
---
[Advent of Code](https://adventofcode.com/2020/about) is an Advent calendar of small programming puzzles for a variety of skill sets and skill levels that can be solved in any programming language you like. 

You don't need a computer science background to participate - just a little programming knowledge and some problem solving skills will get you pretty far. Nor do you need a fancy computer; every problem has a solution that completes in at most 15 seconds on ten-year-old hardware.

This project is my solution to those problems.  

You can view the code in my [advent2020 repository](https://github.com/barrychocolate/Advent2020/).

It is also published using Github Pages which can be viewed at [https://barrychocolate.github.io/Advent2020/solutions.html](https://barrychocolate.github.io/Advent2020/solutions.html)

### Setup
```{r initial_setup}
#Set this option so large numbers are not output in scientific notation
options(scipen=999)
```

## Day 1: Report Repair
Before you leave, the Elves in accounting just need you to fix your expense report (your puzzle input); apparently, something isn't quite adding up.  [Full details....](https://adventofcode.com/2020/day/1)

### Day 1: Load the data
Note participants don't always get the same starting dataset so you should replace the day1.txt dataset with your own if you want to work out your answers.

```{r warning=FALSE, message=FALSE, Day1_load}
#Load the data that will be used for Part 1 & Part 2
#The data can be  downloaded from https://adventofcode.com/2020/day/1/

day1 <- read.csv('data/day1.txt',header = FALSE)

#convert to vector
day1 <- unlist(day1)
```

### Day 1: Part 1
Specifically, they need you to find the two entries that sum to 2020 and then multiply those two numbers together.

```{r day1_part1}
#get all combinations of the numbers
day1_p1_combinations <- expand.grid(day1, day1)

#Calulate the sum of the two numbers so we can check if it equals 2020
day1_p1_combinations$agg <- day1_p1_combinations$Var1 + day1_p1_combinations$Var2

#Remove were the number is paired with itself
day1_p1_combinations <- day1_p1_combinations[day1_p1_combinations$Var1 != day1_p1_combinations$Var2, ]

#Deduplicate
day1_p1_combinations <- day1_p1_combinations[day1_p1_combinations$Var1 < day1_p1_combinations$Var2, ]

#Filter to find those with a sum of 2020
sum_to_2020_p1 <- day1_p1_combinations[day1_p1_combinations$agg==2020, ]

#Multiply the two numbers to get the answer
answer_d1p1 <- sum_to_2020_p1$Var1 * sum_to_2020_p1$Var2
```

The answer to **Day 1 Part 1** is `r answer_d1p1`

### Day 1: Part 2

The Elves in accounting are thankful for your help; one of them even offers you a starfish coin they had left over from a past vacation. They offer you a second one if you can find three numbers in your expense report that meet the same criteria.

```{r day1_part2}
#get all combinations of the numbers
day1_p2_combinations <- expand.grid(day1, day1, day1)

#Calulate the sum of the two numbers so we can check if it equals 2020
day1_p2_combinations$agg <- day1_p2_combinations$Var1 + day1_p2_combinations$Var2 + day1_p2_combinations$Var3

#Remove were the number is paired with itself
day1_p2_combinations <- day1_p2_combinations[day1_p2_combinations$Var1 != day1_p2_combinations$Var2, ]

day1_p2_combinations <- day1_p2_combinations[day1_p2_combinations$Var2 != day1_p2_combinations$Var3, ]

#Deduplicate
day1_p2_combinations <- day1_p2_combinations[day1_p2_combinations$Var1 < day1_p2_combinations$Var2, ]

day1_p2_combinations <- day1_p2_combinations[day1_p2_combinations$Var2 < day1_p2_combinations$Var3, ]

#Filter to find those with a sum of 2020
sum_to_2020_p2 <- day1_p2_combinations[day1_p2_combinations$agg==2020, ]

#Multiply the two numbers to get the answer
#Only using the first record as the second record has the same numbers but alternated
answer_d1p2 <- sum_to_2020_p2[1,]$Var1 * sum_to_2020_p2[1,]$Var2 * sum_to_2020_p2[1,]$Var3
```

The answer to **Day 1 Part 2** is `r answer_d1p2`

## Day 2: Password Philosophy
Your flight departs in a few days from the coastal airport; the easiest way down to the coast from here is via toboggan.

The shopkeeper at the North Pole Toboggan Rental Shop is having a bad day. "Something's wrong with our computers; we can't log in!" You ask if you can take a look.

Their password database seems to be a little corrupted: some of the passwords wouldn't have been allowed by the Official Toboggan Corporate Policy that was in effect when they were chosen.
[Full details....](https://adventofcode.com/2020/day/2)

### Day 2: Load the data
Note participants don't always get the same starting dataset so you should replace the day1.txt dataset with your own if you want to work out your answers.

```{r warning=FALSE, message=FALSE, day2_load}
#Remove previous days data
rm(list = ls())

#these librarys allow the use of the separate function and substr
library(tidyr, stringr)

#Load the data that will be used for Part 1 & Part 2
#The data can be  downloaded from https://adventofcode.com/2020/day/2/

day2 <- read.table('data/day2.txt', sep = " " , header = F, stringsAsFactors= F)
names(day2) <- c("char_count", "criteria", "pword")

#Remove : from criteria column
day2$criteria <- substr(day2$criteria, 1, 1)

#Split min and max and convert to int
day2 <- tidyr::separate(day2, char_count, c('mincount', 'maxcount'), sep = "-", convert = TRUE)


```

### Day 2: Part 1
To try to debug the problem, they have created a list (your puzzle input) of passwords (according to the corrupted database) and the corporate policy when that password was set.

For example, suppose you have the following list:
1-3 a: abcde
1-3 b: cdefg
2-9 c: ccccccccc

Each line gives the password policy and then the password. The password policy indicates the lowest and highest number of times a given letter must appear for the password to be valid. For example, 1-3 a means that the password must contain a at least 1 time and at most 3 times.

In the above example, 2 passwords are valid. The middle password, cdefg, is not; it contains no instances of b, but needs at least 1. The first and third passwords are valid: they contain one a or nine c, both within the limits of their respective policies.

How many passwords are valid according to their policies?

```{r day2_part1}
#Get a count of the number of times the criteria stirng appears in the password
day2$criteria_count <- stringr::str_count(day2$pword, day2$criteria)

#filter the dataset to includ only those where the count is 
day2_p1_valid <- day2 %>%
    dplyr::filter(criteria_count >= mincount & criteria_count <= maxcount)

answer_d2p1 <- nrow(day2_p1_valid)
```

The answer to **Day 2 Part 1** is `r answer_d2p1`

### Day 2: Part 2
While it appears you validated the passwords correctly, they don't seem to be what the Official Toboggan Corporate Authentication System is expecting.

The shopkeeper suddenly realizes that he just accidentally explained the password policy rules from his old job at the sled rental place down the street! The Official Toboggan Corporate Policy actually works a little differently.

Each policy actually describes two positions in the password, where 1 means the first character, 2 means the second character, and so on. (Be careful; Toboggan Corporate Policies have no concept of "index zero"!) Exactly one of these positions must contain the given letter. Other occurrences of the letter are irrelevant for the purposes of policy enforcement.

How many passwords are valid according to the new interpretation of the policies?

```{r day2_part2}
#extract the characters at the positions
day2$charpos1 <-substr(day2$pword, start = day2$mincount, stop = day2$mincount)
day2$charpos2 <-substr(day2$pword, start = day2$maxcount, stop = day2$maxcount)

#Filter toidentify the valid passwords
#i.e the character is in one of the two positions but not both
day2_p2_valid <- day2 %>%
    dplyr::filter( (charpos1 == criteria | charpos2 == criteria) & charpos1 != charpos2) 

#Get a count of the number of valid passowrds
answer_d2p2 <- nrow(day2_p2_valid)
```

The answer to **Day 2 Part 2** is `r answer_d2p2`

## Day 3: Toboggan Trajectory
With the toboggan login problems resolved, you set off toward the airport. While travel by toboggan might be easy, it's certainly not safe: there's very minimal steering and the area is covered in trees. You'll need to see which angles will take you near the fewest trees.

Due to the local geology, trees in this area only grow on exact integer coordinates in a grid. You make a map (your puzzle input) of the open squares (.) and trees (#) you can see.

[Full details....](https://adventofcode.com/2020/day/3)

### Day 3: Load the data
Note participants don't always get the same starting dataset so you should replace the day1.txt dataset with your own if you want to work out your answers.

```{r warning=FALSE, message=FALSE, day3_load}
#Remove previous day's data
rm(list = ls())

#dplyr for filter, numebrs for mod function
library(dplyr, numbers)

#Load the data that will be used for Part 1 & Part 2
#The data can be  downloaded from https://adventofcode.com/2020/day/3/

day3 <- read.csv('data/day3.txt',header = FALSE, stringsAsFactors = FALSE, col.names = c('map'))
```

### Day 3: Part 1
Starting at the top-left corner of your map and following a slope of right 3 and down 1, how many trees would you encounter?

```{r day3_part1}
#create a function to count the number of trees
#I have used a function as Advent of Code is likely going to ask me to do something similar in Part 2

howmanytrees <- function(data, starting_position, rmove, dmove, maprowlength){
#create a temporary copy of the data for use within the function
temp_data<-data

#Remove rows we won't land on
temp_data <- temp_data %>%
    dplyr::filter (numbers::mod(1:nrow(temp_data)-1,dmove)==0)

#Calculate the position on the map where the toboggan will land
temp_data$position <- numbers::mod((starting_position +((1:nrow(temp_data)-1)*rmove)), maprowlength)

#update any position zero to the maprowlength
temp_data$position[temp_data$position ==0] <- maprowlength

#Use the position to calculate what is at that position and store it in the target column
temp_data$target <- substr(temp_data$map, start = temp_data$position, stop = temp_data$position)

#Get a count of all targets that were trees
#Filter rows to get only those that would be targeted based on the downward movement
return(as.numeric(nrow(temp_data %>% dplyr::filter (target == '#'))))
}

#Call the function with the day 3 data, starting at position 1, with a movement of 3 and a map row length of 31
answer_d3p1<-howmanytrees(day3, 1, 3, 1, 31)

```

The answer to **Day 3 Part 1** is `r answer_d3p1`

### Day 3: Part 2
Time to check the rest of the slopes - you need to minimize the probability of a sudden arboreal stop, after all.

Determine the number of trees you would encounter if, for each of the following slopes, you start at the top-left corner and traverse the map all the way to the bottom:

Right 1, down 1.
Right 3, down 1. (This is the slope you already checked.)
Right 5, down 1.
Right 7, down 1.
Right 1, down 2.
In the above example, these slopes would find 2, 7, 3, 4, and 2 tree(s) respectively; multiplied together, these produce the answer 336.

```{r day3_part2}
#Call the function for the Part 2 requirements
r1d1 <- howmanytrees(day3, 1, 1, 1, 31)
r3d1 <- howmanytrees(day3, 1, 3, 1, 31)
r5d1 <- howmanytrees(day3, 1, 5, 1, 31)
r7d1 <- howmanytrees(day3, 1, 7, 1, 31)
r1d2 <- howmanytrees(day3, 1, 1, 2, 31) 

#multiply the results together to get the answer to part 2
answer_d3p2 <- r1d1 * r3d1 * r5d1 * r7d1 * r1d2
```

The answer to **Day 3 Part 2** is `r format(answer_d3p2, nsmall=0, big.mark=",")`


