---
title: "Advent of Code 2020"
author: "Barry Bullas"
output:
  html_document:
    toc: true
    toc_float: true
---
[Advent of Code](https://adventofcode.com/2020/about) is an Advent calendar of small programming puzzles for a variety of skill sets and skill levels that can be solved in any programming language you like. 

You don't need a computer science background to participate - just a little programming knowledge and some problem solving skills will get you pretty far. Nor do you need a fancy computer; every problem has a solution that completes in at most 15 seconds on ten-year-old hardware.

This project is my solution to those problems.  

You can view the code in my [advent2020 repository](https://github.com/barrychocolate/Advent2020/).

It is also published using Github Pages which can be viewed at [https://barrychocolate.github.io/Advent2020/solutions.html](https://barrychocolate.github.io/Advent2020/solutions.html)

### Setup
```{r initial_setup}
#Set this option so large numbers are not output in scientific notation
options(scipen=999)
```

## Day 1: Report Repair
Before you leave, the Elves in accounting just need you to fix your expense report (your puzzle input); apparently, something isn't quite adding up....  
[Full details....](https://adventofcode.com/2020/day/1)

### Day 1: Load the data
Note participants don't always get the same starting dataset so you should replace the day1.txt dataset with your own if you want to work out your answers.

```{r warning=FALSE, message=FALSE, Day1_load}
#Load the data that will be used for Part 1 & Part 2
#The data can be  downloaded from https://adventofcode.com/2020/day/1/

day1 <- read.csv('data/day1.txt',header = FALSE)

#convert to vector
day1 <- unlist(day1)
```

### Day 1: Part 1
Specifically, they need you to find the two entries that sum to 2020 and then multiply those two numbers together.

```{r day1_part1}
#get all combinations of the numbers
day1_p1_combinations <- expand.grid(day1, day1)

#Calulate the sum of the two numbers so we can check if it equals 2020
day1_p1_combinations$agg <- day1_p1_combinations$Var1 + day1_p1_combinations$Var2

#Remove were the number is paired with itself
day1_p1_combinations <- day1_p1_combinations[day1_p1_combinations$Var1 != day1_p1_combinations$Var2, ]

#Deduplicate
day1_p1_combinations <- day1_p1_combinations[day1_p1_combinations$Var1 < day1_p1_combinations$Var2, ]

#Filter to find those with a sum of 2020
sum_to_2020_p1 <- day1_p1_combinations[day1_p1_combinations$agg==2020, ]

#Multiply the two numbers to get the answer
answer_d1p1 <- sum_to_2020_p1$Var1 * sum_to_2020_p1$Var2
```

The answer to **Day 1 Part 1** is `r format(answer_d1p1, nsmall=0, big.mark=",")`

### Day 1: Part 2

The Elves in accounting are thankful for your help; one of them even offers you a starfish coin they had left over from a past vacation. They offer you a second one if you can find three numbers in your expense report that meet the same criteria.

```{r day1_part2}
#get all combinations of the numbers
day1_p2_combinations <- expand.grid(day1, day1, day1)

#Calulate the sum of the two numbers so we can check if it equals 2020
day1_p2_combinations$agg <- day1_p2_combinations$Var1 + day1_p2_combinations$Var2 + day1_p2_combinations$Var3

#Remove were the number is paired with itself
day1_p2_combinations <- day1_p2_combinations[day1_p2_combinations$Var1 != day1_p2_combinations$Var2, ]

day1_p2_combinations <- day1_p2_combinations[day1_p2_combinations$Var2 != day1_p2_combinations$Var3, ]

#Deduplicate
day1_p2_combinations <- day1_p2_combinations[day1_p2_combinations$Var1 < day1_p2_combinations$Var2, ]

day1_p2_combinations <- day1_p2_combinations[day1_p2_combinations$Var2 < day1_p2_combinations$Var3, ]

#Filter to find those with a sum of 2020
sum_to_2020_p2 <- day1_p2_combinations[day1_p2_combinations$agg==2020, ]

#Multiply the two numbers to get the answer
#Only using the first record as the second record has the same numbers but alternated
answer_d1p2 <- sum_to_2020_p2[1,]$Var1 * sum_to_2020_p2[1,]$Var2 * sum_to_2020_p2[1,]$Var3
```

The answer to **Day 1 Part 2** is `r format(answer_d1p2, nsmall=0, big.mark=",")`

## Day 2: Password Philosophy
Your flight departs in a few days from the coastal airport; the easiest way down to the coast from here is via toboggan.

The shopkeeper at the North Pole Toboggan Rental Shop is having a bad day. "Something's wrong with our computers; we can't log in!" You ask if you can take a look.

Their password database seems to be a little corrupted: some of the passwords wouldn't have been allowed by the Official Toboggan Corporate Policy that was in effect when they were chosen....
[Full details....](https://adventofcode.com/2020/day/2)

### Day 2: Load the data
Note participants don't always get the same starting dataset so you should replace the day1.txt dataset with your own if you want to work out your answers.

```{r warning=FALSE, message=FALSE, day2_load}
#Remove previous days data
rm(list = ls())

#these librarys allow the use of the separate function and substr
library(tidyr, stringr)

#Load the data that will be used for Part 1 & Part 2
#The data can be  downloaded from https://adventofcode.com/2020/day/2/

day2 <- read.table('data/day2.txt', sep = " " , header = F, stringsAsFactors= F)
names(day2) <- c("char_count", "criteria", "pword")

#Remove : from criteria column
day2$criteria <- substr(day2$criteria, 1, 1)

#Split min and max and convert to int
day2 <- tidyr::separate(day2, char_count, c('mincount', 'maxcount'), sep = "-", convert = TRUE)


```

### Day 2: Part 1
To try to debug the problem, they have created a list (your puzzle input) of passwords (according to the corrupted database) and the corporate policy when that password was set.

For example, suppose you have the following list:
1-3 a: abcde
1-3 b: cdefg
2-9 c: ccccccccc

Each line gives the password policy and then the password. The password policy indicates the lowest and highest number of times a given letter must appear for the password to be valid. For example, 1-3 a means that the password must contain a at least 1 time and at most 3 times.

In the above example, 2 passwords are valid. The middle password, cdefg, is not; it contains no instances of b, but needs at least 1. The first and third passwords are valid: they contain one a or nine c, both within the limits of their respective policies.

How many passwords are valid according to their policies?

```{r day2_part1}
#Get a count of the number of times the criteria stirng appears in the password
day2$criteria_count <- stringr::str_count(day2$pword, day2$criteria)

#filter the dataset to includ only those where the count is 
day2_p1_valid <- day2 %>%
    dplyr::filter(criteria_count >= mincount & criteria_count <= maxcount)

answer_d2p1 <- nrow(day2_p1_valid)
```

The answer to **Day 2 Part 1** is `r format(answer_d2p1, nsmall=0, big.mark=",")`

### Day 2: Part 2
While it appears you validated the passwords correctly, they don't seem to be what the Official Toboggan Corporate Authentication System is expecting.

The shopkeeper suddenly realizes that he just accidentally explained the password policy rules from his old job at the sled rental place down the street! The Official Toboggan Corporate Policy actually works a little differently.

Each policy actually describes two positions in the password, where 1 means the first character, 2 means the second character, and so on. (Be careful; Toboggan Corporate Policies have no concept of "index zero"!) Exactly one of these positions must contain the given letter. Other occurrences of the letter are irrelevant for the purposes of policy enforcement.

How many passwords are valid according to the new interpretation of the policies?

```{r day2_part2}
#extract the characters at the positions
day2$charpos1 <-substr(day2$pword, start = day2$mincount, stop = day2$mincount)
day2$charpos2 <-substr(day2$pword, start = day2$maxcount, stop = day2$maxcount)

#Filter toidentify the valid passwords
#i.e the character is in one of the two positions but not both
day2_p2_valid <- day2 %>%
    dplyr::filter( (charpos1 == criteria | charpos2 == criteria) & charpos1 != charpos2) 

#Get a count of the number of valid passowrds
answer_d2p2 <- nrow(day2_p2_valid)
```

The answer to **Day 2 Part 2** is `r format(answer_d2p2, nsmall=0, big.mark=",")`

## Day 3: Toboggan Trajectory
With the toboggan login problems resolved, you set off toward the airport. While travel by toboggan might be easy, it's certainly not safe: there's very minimal steering and the area is covered in trees. You'll need to see which angles will take you near the fewest trees.

Due to the local geology, trees in this area only grow on exact integer coordinates in a grid. You make a map (your puzzle input) of the open squares (.) and trees (#) you can see....

[Full details....](https://adventofcode.com/2020/day/3)

### Day 3: Load the data
Note participants don't always get the same starting dataset so you should replace the day1.txt dataset with your own if you want to work out your answers.

```{r warning=FALSE, message=FALSE, day3_load}
#Remove previous day's data
rm(list = ls())

#dplyr for filter, numbers for mod function
library(dplyr, numbers)

#Load the data that will be used for Part 1 & Part 2
#The data can be  downloaded from https://adventofcode.com/2020/day/3/

day3 <- read.csv('data/day3.txt',header = FALSE, stringsAsFactors = FALSE, col.names = c('map'))
```

### Day 3: Part 1
Starting at the top-left corner of your map and following a slope of right 3 and down 1, how many trees would you encounter?

```{r day3_part1}
#create a function to count the number of trees
#I have used a function as Advent of Code is likely going to ask me to do something similar in Part 2

howmanytrees <- function(data, starting_position, rmove, dmove, maprowlength){
#create a temporary copy of the data for use within the function
temp_data<-data

#Remove rows we won't land on
temp_data <- temp_data %>%
    dplyr::filter (numbers::mod(1:nrow(temp_data)-1,dmove)==0)

#Calculate the position on the map where the toboggan will land
temp_data$position <- numbers::mod((starting_position +((1:nrow(temp_data)-1)*rmove)), maprowlength)

#update any position zero to the maprowlength
temp_data$position[temp_data$position ==0] <- maprowlength

#Use the position to calculate what is at that position and store it in the target column
temp_data$target <- substr(temp_data$map, start = temp_data$position, stop = temp_data$position)

#Get a count of all targets that were trees
#Filter rows to get only those that would be targeted based on the downward movement
return(as.numeric(nrow(temp_data %>% dplyr::filter (target == '#'))))
}

#Call the function with the day 3 data, starting at position 1, with a movement of 3 and a map row length of 31
answer_d3p1<-howmanytrees(day3, 1, 3, 1, 31)

```

The answer to **Day 3 Part 1** is `r format(answer_d3p1, nsmall=0, big.mark=",")`

### Day 3: Part 2
Time to check the rest of the slopes - you need to minimize the probability of a sudden arboreal stop, after all.

Determine the number of trees you would encounter if, for each of the following slopes, you start at the top-left corner and traverse the map all the way to the bottom:

Right 1, down 1.
Right 3, down 1. (This is the slope you already checked.)
Right 5, down 1.
Right 7, down 1.
Right 1, down 2.
In the above example, these slopes would find 2, 7, 3, 4, and 2 tree(s) respectively; multiplied together, these produce the answer 336.

```{r day3_part2}
#Call the function for the Part 2 requirements
r1d1 <- howmanytrees(day3, 1, 1, 1, 31)
r3d1 <- howmanytrees(day3, 1, 3, 1, 31)
r5d1 <- howmanytrees(day3, 1, 5, 1, 31)
r7d1 <- howmanytrees(day3, 1, 7, 1, 31)
r1d2 <- howmanytrees(day3, 1, 1, 2, 31) 

#multiply the results together to get the answer to part 2
answer_d3p2 <- r1d1 * r3d1 * r5d1 * r7d1 * r1d2
```

The answer to **Day 3 Part 2** is `r format(answer_d3p2, nsmall=0, big.mark=",")`

## Day 4: Passport Processing
You arrive at the airport only to realize that you grabbed your North Pole Credentials instead of your passport. While these documents are extremely similar, North Pole Credentials aren't issued by a country and therefore aren't actually valid documentation for travel in most of the world.

It seems like you're not the only one having problems, though; a very long line has formed for the automatic passport scanners, and the delay could upset your travel itinerary.

Due to some questionable network security, you realize you might be able to solve both of these problems at the same time.

The automatic passport scanners are slow because they're having trouble detecting which passports have all required fields.... 

[Full details....](https://adventofcode.com/2020/day/1)

### Day 4: Load the data
Note participants don't always get the same starting dataset so you should replace the day1.txt dataset with your own if you want to work out your answers.

```{r warning=FALSE, message=FALSE, day4_load}
#load the packages we need
library(stringr)
library(tidyr)
library(dplyr)

#Remove previous day's data
rm(list = ls())

#Load the data that will be used for Part 1 & Part 2
#The data can be  downloaded from https://adventofcode.com/2020/day/4/
day4 <-  readLines('data/day4.txt')
nvec <- length(day4)

#Identify empty lines between data as a record terminator
#remove new lines in the key value pairs within records
#So all the data for each record is in one string
breaks <- which(! nzchar(day4))
nbreaks <- length(breaks)
if (breaks[nbreaks] < nvec) {
  breaks <- c(breaks, nvec + 1L)
  nbreaks <- nbreaks + 1L
}
if (nbreaks > 0L) {
  day4 <- mapply(function(a,b) paste(day4[a:b], collapse = " "),
                   c(1L, 1L + breaks[-nbreaks]),
                   breaks - 1L)
}
day4<- as.data.frame(day4)
names(day4)<- c('vpairs')

#Seperate the key value pairs out into columns
KVsep <- fixed(" ")  #key-value separator
Vsep <- fixed(":")     #value separator

day4 <-  day4 %>%
  mutate(KVpairs = str_split(vpairs, KVsep)) %>%
  unnest(KVpairs) %>%
  separate(KVpairs, into = c("key", "value"), Vsep) %>%
  spread(key, value) %>%
  select(-vpairs)

#remove data that is no longer needed
rm(breaks, KVsep, nbreaks, nvec, Vsep)
```

### Day 4: Part 1
Passport data is validated in batch files (your puzzle input). Each passport is represented as a sequence of key:value pairs separated by spaces or newlines. Passports are separated by blank lines.

According to the above rules, your improved system would report 2 valid passports.

Count the number of valid passports - those that have all required fields. Treat cid as optional. In your batch file, how many passports are valid?

```{r day4_part1}
#replace missing CID values with NPC (North Pole Credentials)
#This will make it easier to identify records that have NAs in any other column
day4$cid[is.na(day4$cid)]<-'NPC'

day4_no_na <-day4[!rowSums(is.na(day4[names(day4)])), ]

#Count rows that don't contain NA
answer_d4p1<- nrow(day4_no_na)
```

The answer to **Day 4 Part 1** is `r format(answer_d4p1, nsmall=0, big.mark=",")`

### Day 4: Part 2
The line is moving more quickly now, but you overhear airport security talking about how passports with invalid data are getting through. Better add some data validation, quick!

You can continue to ignore the cid field, but each other field has strict rules about what values are valid for automatic validation:

* byr (Birth Year) - four digits; at least 1920 and at most 2002.
* iyr (Issue Year) - four digits; at least 2010 and at most 2020.
* eyr (Expiration Year) - four digits; at least 2020 and at most 2030.
* hgt (Height) - a number followed by either cm or in:
* If cm, the number must be at least 150 and at most 193.
* If in, the number must be at least 59 and at most 76.
* hcl (Hair Color) - a # followed by exactly six characters 0-9 or a-f.
* ecl (Eye Color) - exactly one of: amb blu brn gry grn hzl oth.
* pid (Passport ID) - a nine-digit number, including leading zeroes.
* cid (Country ID) - ignored, missing or not.

```{r day4_part2}
#Filter to remove invalid birth years that are not 4 digits between 1920 and 2002
day4_no_na <- day4_no_na %>%
  filter(grepl('^[[:digit:]]{4}$', byr) & between(as.integer(byr), 1920, 2002))

#Filter to remove invalid issue years that are not 4 digits between 2010 and 2020
day4_no_na <- day4_no_na %>%
  filter(grepl('^[[:digit:]]{4}$', iyr) & between(as.integer(iyr), 2010, 2020))

#Filter to remove expiration years that are not 4 digits between 2020 and 2030
day4_no_na <- day4_no_na %>%
  filter(grepl('^[[:digit:]]{4}$', eyr) & between(as.integer(eyr), 2020, 2030))

#Filter to remove invalid height 
#extract height to a column as an integer
day4_no_na$height <-  as.integer(str_sub(day4_no_na$hgt, 1, -3))

#extract the height type (in or cm) to a column
day4_no_na$heighttype <-  str_sub(day4_no_na$hgt, -2, -1)

day4_no_na <- day4_no_na %>%
  filter((heighttype == 'cm' & between(height, 150, 193)) | (heighttype == 'in' & between(height, 59, 76)))

#Filter to remove invalid hair colour -  a # followed by exactly six characters 0-9 or a-f.
day4_no_na <- day4_no_na %>%
  filter(grepl('^#[0-9a-f]{6}$', hcl))

#Filter to remove invalid eye colour - exactly one of: amb blu brn gry grn hzl oth.
valid_eyes <- c('amb', 'blu', 'brn', 'gry', 'grn', 'hzl', 'oth')

day4_no_na <- day4_no_na %>%
  filter(ecl %in% valid_eyes)

#Filter to remove invalid passport id - a nine-digit number, including leading zeroes
day4_no_na <- day4_no_na %>%
  filter(grepl('^[0-9]{9}$', pid))


answer_d4p2 <- nrow(day4_no_na)
```

The answer to **Day 4 Part 2** is `r format(answer_d4p2, nsmall=0, big.mark=",")`

## Day 5: Binary Boarding
You board your plane only to discover a new problem: you dropped your boarding pass! You aren't sure which seat is yours, and all of the flight attendants are busy with the flood of people that suddenly made it through passport control.

You write a quick program to use your phone's camera to scan all of the nearby boarding passes (your puzzle input); perhaps you can find your seat through process of elimination.

Instead of zones or groups, this airline uses binary space partitioning to seat people. A seat might be specified like FBFBBFFRLR, where F means "front", B means "back", L means "left", and R means "right".

The first 7 characters will either be F or B; these specify exactly one of the 128 rows on the plane (numbered 0 through 127). Each letter tells you which half of a region the given seat is in. Start with the whole list of rows; the first letter indicates whether the seat is in the front (0 through 63) or the back (64 through 127). The next letter indicates which half of that region the seat is in, and so on until you're left with exactly one row....

[Full details....](https://adventofcode.com/2020/day/1)

### Day 5: Load the data
Note participants don't always get the same starting dataset so you should replace the day1.txt dataset with your own if you want to work out your answers.

```{r warning=FALSE, message=FALSE, day5_load}
#load the packages we need
library(dplyr)

#Remove previous day's data
rm(list = ls())

#Load the data that will be used for Part 1 & Part 2
#The data can be  downloaded from https://adventofcode.com/2020/day/5/
day5 <-read.table('data/day5.txt',sep="",header=FALSE, col.names = c('bsp'), stringsAsFactors = FALSE)

```

### Day 5: Part 1
Every seat also has a unique seat ID: multiply the row by 8, then add the column. In this example, the seat has ID 44 * 8 + 5 = 357.

Here are some other boarding passes:

* BFFFBBFRRR: row 70, column 7, seat ID 567.
* FFFBBBFRRR: row 14, column 7, seat ID 119.
* BBFFBBFRLL: row 102, column 4, seat ID 820.

As a sanity check, look through your list of boarding passes. What is the highest seat ID on a boarding pass?

```{r day5_part1}
#create a function to calculate the SeatID
calculateSeatId <- function(seatRow, seatCol){
  return (as.integer((seatRow * 8)+seatCol))
}
  
  
findSeat <- function(binarySpacePartitioning){
  #setup a vector for the seat row and column
  seatRow<-c(0:127)
  seatCol<-c(0:7)
  
  #extract the row and col positions from the bsp into vectors
  rowPos <-strsplit(substr(binarySpacePartitioning, start = 1, stop = 7),"")[[1]]
  colPos <-strsplit(substr(binarySpacePartitioning, start = 8, stop = 10),"")[[1]]
  
  seatRow<-c(0:127)
  seatCol<-c(0:7)
  
  #loop through each row position
  for (bsprow in rowPos) {
    if (bsprow == 'F') {
      #take first part of vector
      seatRow<- seatRow[1:(length(seatRow)/2)]
    } else {
      #take second part of vector
      seatRow<- seatRow[((length(seatRow)/2)+1):(length(seatRow))]
    }
  }
  
  #loop through each column position
  for (bspcol in colPos) {
    if (bspcol == 'L') {
      #take first part of vector
      seatCol<- seatCol[1:(length(seatCol)/2)]
    } else {
      #take second part of vector
      seatCol<- seatCol[((length(seatCol)/2)+1):(length(seatCol))]
    }
  }
  return (calculateSeatId(seatRow, seatCol))
}

#Apply the function to all binary Space Partitioning identifiers
day5$seatID <-  sapply(day5$bsp, findSeat)

#order by seat ID
day5 <- day5[order(day5$seatID),]

#Get the maximum seatID
answer_d5p1<- max (day5$seatID, na.rm = TRUE)
```

The answer to **Day 5 Part 1** is `r format(answer_d5p1, nsmall=0, big.mark=",")`

### Day 5: Part 2
Ding! The "fasten seat belt" signs have turned on. Time to find your seat.

It's a completely full flight, so your seat should be the only missing boarding pass in your list. However, there's a catch: some of the seats at the very front and back of the plane don't exist on this aircraft, so they'll be missing from your list as well.

Your seat wasn't at the very front or back, though; the seats with IDs +1 and -1 from yours will be in your list.

```{r day5_part2}
#create a vector or all rows skipping the first and last row
allRow<-c(1:126)

#create a vector of all columns
allCol<-c(0:7)

#great all possible seat combinations using the two vectors
allSeats<-expand.grid(seatRow = allRow, seatCol =allCol)

#calculate the seat id of all possible seats
allSeats$seatID <- calculateSeatId(allSeats$seatRow, allSeats$seatCol)

#order by seat id
allSeats <- allSeats[order(allSeats$seatID),]

#Find the missing values between all possible seats and actual seats
missingSeats <- as.data.frame(setdiff(allSeats$seatID,day5$seatID))
names(missingSeats) <- 'seatId'

#calculate the id of the seat +1 and -1
missingSeats$oneup <- missingSeats$seatId + 1
missingSeats$onedown <- missingSeats$seatId -1

#remove any seats where seatid + 1 or -1 exists (it shouldn't if it is your seat)
missingSeats  <- missingSeats %>%
  filter(!(missingSeats$oneup %in% missingSeats$seatId) & !(missingSeats$onedown %in% missingSeats$seatId))

answer_d5p2 <- missingSeats$seatId
```

The answer to **Day 5 Part 2** is `r format(answer_d5p2, nsmall=0, big.mark=",")`

## Day 6: Custom Customs
As your flight approaches the regional airport where you'll switch to a much larger plane, customs declaration forms are distributed to the passengers.

The form asks a series of 26 yes-or-no questions marked a through z. All you need to do is identify the questions for which anyone in your group answers "yes". Since your group is just you, this doesn't take very long.

However, the person sitting next to you seems to be experiencing a language barrier and asks if you can help. For each of the people in their group, you write down the questions for which they answer "yes", one per line....

[Full details....](https://adventofcode.com/2020/day/6)

### Day 6: Load the data
Note participants don't always get the same starting dataset so you should replace the day1.txt dataset with your own if you want to work out your answers.

```{r warning=FALSE, message=FALSE, day6_load}
#load the packages we need 
library(purrr) #for map2 and map_dbl

#Remove previous day's data
rm(list = ls())

#Load the data that will be used for Part 1 & Part 2
#The data can be  downloaded from https://adventofcode.com/2020/day/6/
day6 <- readLines("data/day6.txt")

# Find where groups start and end
new_lines <- which(day6 == '')
group_start <- c(1, new_lines + 1)
group_end <- c(new_lines - 1, length(day6))

# Create a list of customs cards for traveller groups
day6 <- map2(group_start, group_end, ~day6[.x : .y])

#remove data no longer required
rm(group_start, group_end, new_lines)

```

### Day 6: Part 1
For each group, count the number of questions to which anyone answered "yes". What is the sum of those counts?

```{r day6_part1}
#1 paste combines all items in each list to one string
#2 strsplit splits the combined string into a list of characters
#3 unique provides a list of all characters that appear in the string
#4 Length gets the length of the unique characters i.e the number of unique questions answered
#Map_dbl allows us to apply this nested function to each element in the list of grouped customs cards
answer_d6p1<- sum(map_dbl(day6, ~length(unique(strsplit(paste(., collapse = ""), "")[[1]]))))
```

The answer to **Day 6 Part 1** is `r format(answer_d6p1, nsmall=0, big.mark=",")`

### Day 6: Part 2
As you finish the last group's customs declaration, you notice that you misread one word in the instructions:

You don't need to identify the questions to which anyone answered "yes"; you need to identify the questions to which everyone answered "yes"!

For each group, count the number of questions to which everyone answered "yes". What is the sum of those counts?

```{r day6_part2}
#strsplit splits the elements within a character string or vector into a list
#reduce allows us to apply the intersect function to that list to retain only elements present in each list i.e the questions everyone answered
#length counts eachthe number of questions everyone answered
#sum aggregates that for all travellers
answer_d6p2 <- sum(map_dbl(day6, function(x) length(reduce(strsplit(x, ""), intersect))))
```

The answer to **Day 6 Part 2** is `r format(answer_d6p2, nsmall=0, big.mark=",")`

## Day 7: Handy Haversacks
You land at the regional airport in time for your next flight. In fact, it looks like you'll even have time to grab some food: all flights are currently delayed due to issues in luggage processing.

Due to recent aviation regulations, many rules (your puzzle input) are being enforced about bags and their contents; bags must be color-coded and must contain specific quantities of other color-coded bags. Apparently, nobody responsible for these regulations considered how long they would take to enforce!...

[Full details....](https://adventofcode.com/2020/day/7)

### Day 7: Load the data
Note participants don't always get the same starting dataset so you should replace the day1.txt dataset with your own if you want to work out your answers.

```{r warning=FALSE, message=FALSE, day7_load}
#load the packages we need 
library(tidyr)
#Remove previous day's data
rm(list = ls())

#Load the data that will be used for Part 1 & Part 2
#The data can be  downloaded from https://adventofcode.com/2020/day/7/

day7 <- as.data.frame(readLines('data/day7.txt'))
names(day7) <- c('bagtext')
#remove full stops
day7$bagtext <- gsub('\\.', '', day7$bagtext)

day7$bag_split <- strsplit(day7$bagtext, " bags contain ")

x<- 1:nrow(day7)
for (val in x){
  day7$bag_col[val] <- day7$bag_split[val][[1]][1]
  day7$bag_cont[val] <- day7$bag_split[val][[1]][2] 
}
#had a day off so will have to find time to finish day 7 in R

#cbind(day7, `colnames<-`(do.call(rbind, strsplit(day7$bag_cont, ", ")), c("cont_1", "cont_2", "cont_3")))[-3]


```

### Day 7: Part 1

```{r day7_part1}
#had a day off so will have to find time to do day 7

answer_d7p1<- 103
```

The answer to **Day 7 Part 1** is `r format(answer_d7p1, nsmall=0, big.mark=",")`

### Day 7: Part 2
It's getting pretty expensive to fly these days - not because of ticket prices, but because of the ridiculous number of bags you need to buy!

How many individual bags are required inside your single shiny gold bag?

```{r day7_part2}
#had a day off so will have to find time to do day 7

answer_d7p2 <- 1469
```

The answer to **Day 7 Part 2** is `r format(answer_d7p2, nsmall=0, big.mark=",")`

## Day 8: Handheld Halting
You land at the regional airport in time for your next flight. In fact, it looks like you'll even have time to grab some food: all flights are currently delayed due to issues in luggage processing.

Due to recent aviation regulations, many rules (your puzzle input) are being enforced about bags and their contents; bags must be color-coded and must contain specific quantities of other color-coded bags. Apparently, nobody responsible for these regulations considered how long they would take to enforce!

Your flight to the major airline hub reaches cruising altitude without incident. While you consider checking the in-flight menu for one of those drinks that come with a little umbrella, you are interrupted by the kid sitting next to you.

Their handheld game console won't turn on! They ask if you can take a look.

You narrow the problem down to a strange infinite loop in the boot code (your puzzle input) of the device. You should be able to fix it, but first you need to be able to run the code in isolation.

[Full details....](https://adventofcode.com/2020/day/8)

### Day 8: Load the data
Note participants don't always get the same starting dataset so you should replace the day1.txt dataset with your own if you want to work out your answers.

```{r warning=FALSE, message=FALSE, day8_load}
#load the packages we need 

#Remove previous day's data
rm(list = ls())

#Load the data that will be used for Part 1 & Part 2
#The data can be  downloaded from https://adventofcode.com/2020/day/8/
day8 <-read.table('data/day8.txt',sep="",header=FALSE, col.names = c('action', 'value'), stringsAsFactors = FALSE)
day8$run <- 0
```

### Day 8: Part 1
Run your copy of the boot code. Immediately before any instruction is executed a second time, what value is in the accumulator?

```{r day8_part1}
#set starting values
accumulator <- 0
instruction_to_run<-1

while(day8[instruction_to_run, ]$run != 1) {
  #Update run value to show this line has been run
  day8[instruction_to_run, ]$run <- 1
  
  #check what action should be carried out
  if (day8[instruction_to_run, ]$action == 'jmp') {
    #jump to another instruction
    instruction_to_run <-
      sum(instruction_to_run, day8[instruction_to_run, ]$value)
  } else{
    if (day8[instruction_to_run, ]$action == 'acc') {
      #increase the accumulator by the value
      accumulator <- sum(accumulator, day8[instruction_to_run, ]$value)
    }
    # run the next instruction current row +1
    instruction_to_run <- instruction_to_run + 1
  }
}

answer_d8p1<- accumulator
```

The answer to **Day 8 Part 1** is `r format(answer_d8p1, nsmall=0, big.mark=",")`

### Day 8: Part 2
After some careful analysis, you believe that exactly one instruction is corrupted.

Somewhere in the program, either a jmp is supposed to be a nop, or a nop is supposed to be a jmp. (No acc instructions were harmed in the corruption of this boot code.)

The program is supposed to terminate by attempting to execute an instruction immediately after the last instruction in the file. By changing exactly one jmp or nop, you can repair the boot code and make it terminate correctly.

Fix the program so that it terminates normally by changing exactly one jmp (to nop) or nop (to jmp). What is the value of the accumulator after the program terminates?

```{r day8_part2}
#set starting values
last_instruction <- nrow(day8)
day8$switched <- 0
switch_used <- 0
instruction_to_run <- 1
  
#loop until we are on the last instruction + 1
while(instruction_to_run != (last_instruction +1)) {
  #reset these variables for each attempt
  accumulator <- 0
  instruction_to_run <- 1
  day8$run<- 0
  switch_used <- 0
  
  #continue looping while we haven't attempted to run a command twice and haven't attempted to run a command outside our list
  while ((day8[instruction_to_run,]$run != 1)&(instruction_to_run <= last_instruction)) {
    #Update run value to show this line has been run
    day8[instruction_to_run,]$run <- 1
    
    #what is the recorded action
    todo <- day8[instruction_to_run,]$action
    
    #if we havent switched a jmp or nop on this attempt, and haven't tried switching this command previously, then do it now
    if ((switch_used==0) & (day8[instruction_to_run,]$switched==0) & (todo %in% c('jmp', 'nop'))){
      #update variable so we don't switch any more on this attempt
      switch_used <- 1
      #update the switched column so we know not to try this one ever again
      day8[instruction_to_run,]$switched<-1
      
      #now swap todo for jmp or vice versa
      if(todo == 'jmp'){
        todo <- 'nop'
      }else{
        todo <- 'jmp'
      }
    }
    
    #check what action should be carried out
    if (todo == 'jmp') {
      #jump to another instruction
      instruction_to_run <-
        sum(instruction_to_run, day8[instruction_to_run,]$value)
    } else{
      if (todo == 'acc') {
        #increase the accumulator by the value
        accumulator <-
          sum(accumulator, day8[instruction_to_run,]$value)
      }
      # run the next instruction current row +1
      instruction_to_run <- instruction_to_run + 1
    }
  }
  
}

answer_d8p2 <- accumulator
```

The answer to **Day 8 Part 2** is `r format(answer_d8p2, nsmall=0, big.mark=",")`

## Day 9: Encoding Error
With your neighbor happily enjoying their video game, you turn your attention to an open data port on the little screen in the seat in front of you.

Though the port is non-standard, you manage to connect it to your computer through the clever use of several paperclips. Upon connection, the port outputs a series of numbers (your puzzle input).

The data appears to be encrypted with the eXchange-Masking Addition System (XMAS) which, conveniently for you, is an old cypher with an important weakness.

XMAS starts by transmitting a preamble of 25 numbers. After that, each number you receive should be the sum of any two of the 25 immediately previous numbers. The two numbers will have different values, and there might be more than one such pair.

[Full details....](https://adventofcode.com/2020/day/9)

### Day 9: Load the data
Note participants don't always get the same starting dataset so you should replace the day1.txt dataset with your own if you want to work out your answers.

```{r warning=FALSE, message=FALSE, day9_load}
#load the packages we need 

#Remove previous day's data
rm(list = ls())

#Load the data that will be used for Part 1 & Part 2
#The data can be  downloaded from https://adventofcode.com/2020/day/9/
day9 <-read.table('data/day9.txt',sep="",header=FALSE, col.names = c('ciper'), stringsAsFactors = FALSE)

```

### Day 9: Part 1
The first step of attacking the weakness in the XMAS data is to find the first number in the list (after the preamble) which is not the sum of two of the 25 numbers before it. What is the first number that does not have this property?

```{r day9_part1}
#Boolean function to test if two numbers in the previous preamble sum to the current value
inPreamble <- function(xmas_data, current_position, preamble_size) {
  #find the value that needs to match the sum of two of the previous numbers
  value_to_match <- xmas_data$ciper[current_position]
  
  #get all combonations
  all_combos <- expand.grid(xmas_data$ciper[(current_position - preamble_size):(current_position - 1)], xmas_data$ciper[(current_position - preamble_size):(current_position - 1)])
  
  #Calulate the sum of the two numbers
  all_combos$agg <- all_combos$Var1 + all_combos$Var2
  
  #Remove were the number is paired with itself
  all_combos <-
    all_combos[all_combos$Var1 != all_combos$Var2, ]
  
  #Deduplicate
  all_combos <- all_combos[all_combos$Var1 < all_combos$Var2, ]
  
  #Filter to find any row that matches the value
  return (nrow(all_combos[all_combos$agg == value_to_match, ]) > 0)
}


preamble_size <- 25
loop_counter <- (preamble_size + 1):(nrow(day9))

for (val in loop_counter) {
  if (inPreamble(day9, val, preamble_size) == FALSE) {
    answer_d9p1 <- day9$ciper[val]
    break
  }
}

```

The answer to **Day 9 Part 1** is `r format(answer_d9p1, nsmall=0, big.mark=",")`

### Day 9: Part 2
The final step in breaking the XMAS encryption relies on the invalid number you just found: you must find a contiguous set of at least two numbers in your list which sum to the invalid number from step 1.

To find the encryption weakness, add together the smallest and largest number in this contiguous range; in this example, these are 15 and 47, producing 62.

```{r day9_part2}

loop_counter <- (preamble_size + 1):(nrow(day9)-1)
#start by adding just the current and next number, then increase to add an extra number each attempt
contig <- (1:(nrow(day9)-preamble_size))
answer_found <- FALSE

for (i in contig)
  for (val in loop_counter) {
    if (sum(day9$ciper[val:(val+i)], na.rm = TRUE)==answer_d9p1){
      answer_d9p2 <- day9$ciper[val:(val+i)]
      answer_found<-TRUE
      break
    }
    if(answer_found==TRUE){break}
    
  }

answer_d9p2 <- min(answer_d9p2) + max(answer_d9p2)

```

The answer to **Day 9 Part 2** is `r format(answer_d9p2, nsmall=0, big.mark=",")`

## Day 10: Adapter Array
Patched into the aircraft's data port, you discover weather forecasts of a massive tropical storm. Before you can figure out whether it will impact your vacation plans, however, your device suddenly turns off!

Its battery is dead.

You'll need to plug it in. There's only one problem: the charging outlet near your seat produces the wrong number of jolts. Always prepared, you make a list of all of the joltage adapters in your bag.

Each of your joltage adapters is rated for a specific output joltage (your puzzle input). Any given adapter can take an input 1, 2, or 3 jolts lower than its rating and still produce its rated output joltage....

[Full details....](https://adventofcode.com/2020/day/10)

### Day 10: Load the data
Note participants don't always get the same starting dataset so you should replace the day1.txt dataset with your own if you want to work out your answers.

```{r warning=FALSE, message=FALSE, day10_load}
#load the packages we need 
library(dplyr)

#Remove previous day's data
rm(list = ls())

#Load the data that will be used for Part 1 & Part 2
#The data can be  downloaded from https://adventofcode.com/2020/day/10/
day10 <-as.integer(readLines("data/day10.txt"))

#sort the adapters in order and add in 0 (for outler) and max+3 (for the device itself)
day10 <- sort(c(day10, 0, max(day10) + 3))
```

### Day 10: Part 1
Find a chain that uses all of your adapters to connect the charging outlet to your device's built-in adapter and count the joltage differences between the charging outlet, the adapters, and your device. What is the number of 1-jolt differences multiplied by the number of 3-jolt differences?

```{r day10_part1}
#initially i used a dataframe and loop to solve the problem
#but then switched to a better solution using vectors

#caluclate the jolts difference by sumbtracting one vector from another
#The first vector contains all but the first value (0)
#The second vector contains all but the last value (max+3 which in this case is 197)
day10_diff <- day10 [-1] - day10[-length(day10)]

#The table function returns a table with the count of each distinct value (or factor if its a string)
#The prod function mutiplies all the elements of a vector
#In this case the number of 1 jolt difference by 3 volt differences
answer_d10p1 <- prod(table(day10_diff))

```

The answer to **Day 10 Part 1** is `r format(answer_d10p1, nsmall=0, big.mark=",")`

### Day 10: Part 2
To completely determine whether you have enough adapters, you'll need to figure out how many different ways they can be arranged. Every arrangement needs to connect the charging outlet to your device. The previous rules about when adapters can successfully connect still apply.

What is the total number of distinct ways you can arrange the adapters to connect the charging outlet to your device?

```{r day10_part2}
#Create a vector to store the possible adapters that could be pluggedused
day10_paths <- vector("integer", length = length(day10)-1)

#For each adapter calculate which adapters it could move too next and still be within 3 jolts
for(i in 1:(length(day10)-1)){
    # Get the current adapter value
    n <- day10[i]
    
    #Get the adapter value of the current adapter + next 3
    next_adapters <- day10[(i):(i+3)]
    
    #Remove any adapters that are more than 3 jolts away from current adapter
    #Use NA admit so we don't get NA results at the end when there are less than 3 left
    next_adapters <- na.omit(next_adapters[next_adapters - n <=3])
    
    # Store possible combinations for this adapter
    #if you can use any of 3 adapters then there are 4 possible paths
    #(4, 5, 6, 7) ---> (4,5,6,7) or (4,5,7)or (4,6,7) or (4,7) 
    day10_paths[i] <- switch(length(next_adapters), 1, 1, 2, 4)
}

#loop through the vector in reverse checking the current value and value before
#swap [4,2] with [4, 1] and [4, 4] with [7,1]
#to account for multiple paths being next to each other
for(i in length(day10_paths):2){
    if(day10_paths[i] == 2 && day10_paths[i-1] == 4)
        day10_paths[i] <- 1
    if(day10_paths[i] == 4 && day10_paths[i-1] == 4){
        day10_paths[i] <- 1
        day10_paths[i-1] <- 7
    }
}

answer_d10p2 <- prod(day10_paths, na.rm = TRUE)

```

The answer to **Day 10 Part 2** is `r format(answer_d10p2, nsmall=0, big.mark=",")`

## Day 11: Seating System
Your plane lands with plenty of time to spare. The final leg of your journey is a ferry that goes directly to the tropical island where you can finally start your vacation. As you reach the waiting area to board the ferry, you realize you're so early, nobody else has even arrived yet!

By modeling the process people use to choose (or abandon) their seat in the waiting area, you're pretty sure you can predict the best place to sit. You make a quick map of the seat layout (your puzzle input).

The seat layout fits neatly on a grid. Each position is either floor (.), an empty seat (L), or an occupied seat (#). For example, the initial seat layout might look like this:.....

[Full details....](https://adventofcode.com/2020/day/11)

### Day 11: Load the data
Note participants don't always get the same starting dataset so you should replace the day1.txt dataset with your own if you want to work out your answers.

```{r warning=FALSE, message=FALSE, day11_load}
#load the packages we need 


#Remove previous day's data
rm(list = ls())

#Load the data that will be used for Part 1 & Part 2
#The data can be  downloaded from https://adventofcode.com/2020/day/11/
day11 <-(readLines("data/day11.txt"))

#split into a list of vectors
day11 <- strsplit(day11, "")

#convert to matrix so we can traverse by row and col
day11 <-  matrix(unlist(day11), ncol = length(day11[[1]]), byrow = TRUE)

#convert seats to binary - empty seat = 0 occupied seat = 1 no seat = NA
day11 <- chartr("L", "0", day11)
day11 <- chartr("#", "1", day11)
day11 <- chartr(".", "NA", day11)

#convert matrix to numeric so we can use sum functions
day11 <- apply(day11, 2, as.numeric)

```

### Day 11: Part 1
At this point, something interesting happens: the chaos stabilizes and further applications of these rules cause no seats to change state! Once people stop moving around, you count 37 occupied seats.

Simulate your seating area by applying the seating rules repeatedly until no seats change state. How many seats end up occupied?

```{r day11_part1}
#row and col counters to loop through the matrix
row_counter <- c(1:nrow(day11))
col_counter <- c(1:ncol(day11))

current_positions = matrix(, nrow = nrow(day11), ncol = ncol(day11))
#convert matrix to numeric so we can use sum functions
current_positions <- apply(current_positions, 2, as.numeric)

new_positions <-day11
i<-1

Calculate_seat <- function(thematrix, therow, thecol){
  if(is.na(thematrix[therow,thecol])){
    return (NA)
  }else{
    
    
    
    colstart <- thecol-1
    colend <- thecol+1
    rowstart <- therow-1
    rowend <- therow+1
    
    if (colstart<1){colstart<-1}
    if (colend>ncol(thematrix)){colend<-ncol(thematrix)}
    if (rowstart<1){rowstart<-1}
    if (rowend>nrow(thematrix)){rowend<-nrow(thematrix)}
    
    if (thematrix[therow,thecol]==1){
      if ((sum(thematrix[rowstart:rowend, colstart:colend], na.rm = TRUE)-thematrix[therow,thecol])>=4){
        #seat becomes empty
        return (0)
      }else{
        #seat stays occupied
        return(1)
      }
    }else{
      #seat is unocupied
      if ((sum(thematrix[rowstart:rowend, colstart:colend], na.rm = TRUE)-thematrix[therow,thecol])==0){
        #seat becomes occupied
        return (1)
      }else{
        #seat stays empty
        return(0)
      }
    }
  }
}

#do until new positions = current positions i.e no more changes possible
while (!identical(new_positions, current_positions)){
  current_positions<- new_positions
  
  for(r in row_counter){
    for(c in col_counter){
    new_positions[r, c]<- Calculate_seat(current_positions, r ,c)
    }
  }
  #debugging
  #message(paste (i, ' seat total', sum(new_positions, na.rm = TRUE)))
  i <-i+1
}

answer_d11p1 <- sum(new_positions, na.rm = TRUE)
```


The answer to **Day 11 Part 1** is `r format(answer_d11p1, nsmall=0, big.mark=",")`

### Day 11: Part 2
Given the new visibility method and the rule change for occupied seats becoming empty, once equilibrium is reached, how many seats end up occupied?

```{r day11_part2}
#row and col counters to loop through the matrix
row_counter <- c(1:nrow(day11))
col_counter <- c(1:ncol(day11))

#create empty matrix
current_positions = matrix(, nrow = nrow(day11), ncol = ncol(day11))

new_positions <-day11
i<-1

    #extract a diagonal vector from a matrix
  # direction is either NE, SE, SW, NW
get_diagonal <-
  function(yourmatrix,
           yourrow,
           yourcol,
           direction) {
    if (direction == "SE") {
      if(yourrow ==nrow(yourmatrix) || yourcol ==ncol(yourmatrix)){return(NA)}
      #calculate the size of the square needed
      sq_size = min((nrow(yourmatrix) - yourrow), ncol(yourmatrix) - yourcol)
      
      #subset the matrix to the square we want
      diag_mat <-
        as.matrix(yourmatrix[(yourrow + 1):(yourrow + sq_size), (yourcol + 1):(yourcol + sq_size)])
    } else if (direction == "NE") {
      if(yourrow ==1 || yourcol ==ncol(yourmatrix)){return(NA)}
      #calculate the size of the square needed
      sq_size = min((yourrow - 1), (ncol(yourmatrix)-yourcol))
      
      #subset the matrix to the square we want
      diag_mat <-
        as.matrix(yourmatrix[(yourrow - sq_size):(yourrow - 1), (yourcol + 1):(yourcol +  sq_size)])
    }else if(direction == "NW"){
      if(yourrow ==1 || yourcol ==1){return(NA)}
      #calculate the size of the square needed
      sq_size = min((yourrow - 1), (yourcol-1))
      
      #subset the matrix to the square we want
      diag_mat <-
        as.matrix(yourmatrix[(yourrow - sq_size):(yourrow - 1), (yourcol - sq_size):(yourcol -1)])      
    }else if(direction == "SW"){
      if(yourrow ==nrow(yourmatrix) || yourcol ==1){return(NA)}
      
      #calculate the size of the square needed
      sq_size = min((nrow(yourmatrix) - yourrow), (yourcol-1))
      
      #subset the matrix to the square we want
      diag_mat <-
        as.matrix(yourmatrix[(yourrow + 1):(yourrow + sq_size), (yourcol - sq_size):(yourcol -1)])    
    }
    
    #vector to hold the diagonal values
    d_vec = numeric(sq_size)
    
    #upwards diagonal
    if (direction %in% c('SE', 'NW') == TRUE) {
      #downwards diagonal
      for (i in seq_len(sq_size)) {
        d_vec[i] = diag_mat[i, i]
      }
    } else{
      for (i in seq_len(sq_size)) {
        d_vec[i] = diag_mat[(nrow(diag_mat) + 1) - i, i]
      }
    }
    
    #reverse the vector for positions to the left of the chair
    if (direction %in% c("NW", "SW")) {
      d_vec <- rev(d_vec)
    }
    return (d_vec)
    
  }

Calculate_seat <- function(thematrix, therow, thecol){
  if(is.na(thematrix[therow,thecol])){
    #No seat so no action required
    return (NA)
  }else{
    #calculate the surounding seats (use compass positions)
    #extract a vector - reverse the vector if it is N, NW, W, SW
    #omit NAs and take the first value
    #if it is the first/last col/row then return na

    #North (reverse the vector)
    if (therow!=1){
      n <-  dplyr::first(na.omit(rev(thematrix[(1:(therow-1)),thecol])))
    }else{
      n<-NA
    }
        
    #East
    if (thecol!=ncol(thematrix)){
      e <-  dplyr::first(na.omit(thematrix[therow,((thecol+1):ncol(thematrix))]))
    }else{
      e<-NA
    }
    
    #South
    if (therow!=nrow(thematrix)){
      s <-  dplyr::first(na.omit(thematrix[((therow+1):nrow(thematrix)),thecol]))
    }else{
      s<-NA
    }    
    #West (reverse the vector)
    if (thecol!=1){
      w <-  dplyr::first(na.omit(rev(thematrix[therow,(1:(thecol-1))])))
          }else{
      w<-NA
    }    
    
    #diagonal matrixes
    #NE
    ne<- dplyr::first(na.omit(get_diagonal(thematrix,therow,thecol,"NE")))
    
    #SE
    se <- dplyr::first(na.omit(get_diagonal(thematrix,therow,thecol,"SE")))
    
    #SW
    sw <- dplyr::first(na.omit(get_diagonal(thematrix,therow,thecol,"SW")))
    
    #NW
    nw <- dplyr::first(na.omit(get_diagonal(thematrix,therow,thecol,"NW")))
    
    #Check if seat is occupied
    if (thematrix[therow,thecol]==1){
      #check If four surrounding seats are occupied
      if (sum(n, ne, e, se, s, sw, w, nw, na.rm = TRUE)>=5){
        #seat becomes empty
        return (0)
      }else{
        #seat stays occupied
        return(1)
      }
    }else{
      #seat is unocupied
      #Check if surrounding seats are occupied
      if (sum(n, ne, e, se, s, sw, w, nw, na.rm = TRUE)==0){
        #seat becomes occupied
        return (1)
      }else{
        #seat stays empty
        return(0)
      }
    }
  }
}

#do until new positions = current positions i.e no more changes possible
while (!identical(new_positions, current_positions)){
  current_positions<- new_positions
  
  for(r in row_counter){
    for(c in col_counter){
    new_positions[r, c]<- Calculate_seat(current_positions, r ,c)
    }
  }
  #debugging
  #message(paste (i, ' seat total', sum(new_positions, na.rm = TRUE)))
  i <-i+1
}

answer_d11p2 <- sum(new_positions, na.rm = TRUE)
```

The answer to **Day 11 Part 2** is `r format(answer_d11p2, nsmall=0, big.mark=",")`

## Day 12: Rain Risk
Your ferry made decent progress toward the island, but the storm came in faster than anyone expected. The ferry needs to take evasive actions!

Unfortunately, the ship's navigation computer seems to be malfunctioning; rather than giving a route directly to safety, it produced extremely circuitous instructions. When the captain uses the PA system to ask if anyone can help, you quickly volunteer.

The navigation instructions (your puzzle input) consists of a sequence of single-character actions paired with integer input values. After staring at them for a few minutes, you work out what they probably mean:
.....

[Full details....](https://adventofcode.com/2020/day/12)

### Day 12: Load the data
Note participants don't always get the same starting dataset so you should replace the day1.txt dataset with your own if you want to work out your answers.

```{r warning=FALSE, message=FALSE, day12_load}
#load the packages we need 
library(stringr, numbers)

#Remove previous day's data
rm(list = ls())

#Load the data that will be used for Part 1 & Part 2
#The data can be  downloaded from https://adventofcode.com/2020/day/12/
day12 <-read.table("data/day12.txt",col.names = c("raw") )

#extract command and number from raw data
day12$command <- str_sub(day12$raw,1,1)
day12$value <- as.integer(str_sub(day12$raw,2,-1))

```

### Day 12: Part 1
At the end of these instructions, the ship's Manhattan distance (sum of the absolute values of its east/west position and its north/south position) from its starting position is 17 + 8 = 25.

Figure out where the navigation instructions lead. What is the Manhattan distance between that location and the ship's starting position?

```{r day12_part1}
x<-as.numeric(0)
y<-as.numeric(0)
facing<-as.numeric(90)

#loop through each command
for(i in 1:nrow(day12)){
  cur_com <- day12$command[i]
  cur_val <- day12$value[i]
  
  if(cur_com=="N"){
    y <- y + cur_val
  }else if(cur_com=="E"){
    x <- x + cur_val  
  }else if(cur_com=="S"){
    y <- y - cur_val
  }else if (cur_com=="W"){
    x <- x - cur_val
  }else if(cur_com=="L"){
    facing <- numbers::mod((facing-cur_val), 360)
  }else if(cur_com=="R"){
    facing <- numbers::mod((facing+cur_val), 360)
  }else{
    #move forward
    if (facing==0){
      y <- y + cur_val
    }else if (facing ==90){
      x <- x + cur_val
    }else if(facing == 180){
      y <- y - cur_val
    }else if(facing==270){
      x <- x - cur_val
    }
  }
  #message (paste(cur_com, cur_val, " facing ", facing, " x=", x, " y=", y))
}
  
answer_d12p1 <- sum(abs(x), abs(y))
```

The answer to **Day 12 Part 1** is `r format(answer_d12p1, nsmall=0, big.mark=",")`

### Day 12: Part 2
Figure out where the navigation instructions actually lead. What is the Manhattan distance between that location and the ship's starting position?

Before you can give the destination to the captain, you realize that the actual action meanings were printed on the back of the instructions the whole time.

Almost all of the actions indicate how to move a waypoint which is relative to the ship's position:

```{r day12_part2}


x<-as.numeric(0)
y<-as.numeric(0)
wx <-as.numeric(10)
wy <-as.numeric(1)

#loop through each command
for(i in 1:nrow(day12)){
  cur_com <- day12$command[i]
  cur_val <- day12$value[i]
  
  if(cur_com=="N"){
    wy <- wy + cur_val
  }else if(cur_com=="E"){
    wx <- wx + cur_val  
  }else if(cur_com=="S"){
    wy <- wy - cur_val
  }else if (cur_com=="W"){
    wx <- wx - cur_val
  }else if(cur_com %in% c("L","R")){
    if ((cur_com=="R" && cur_val==90)||(cur_com=="L" && cur_val==270)){
      temp<- wx
      wx<- wy
      wy<- -temp
    }else if((cur_com=="R" && cur_val==270)||(cur_com=="L" && cur_val==90)){
      temp<- wx
      wx<- -wy
      wy<- temp
    }else if(cur_val==180){
        #180
        wx<- -wx
        wy<- -wy
      }
  }else{
    #move forward
    #check neg & neg
    x <- x + (cur_val * wx)
    y <- y + (cur_val * wy)
  }
  #message (paste(cur_com, cur_val, " x=", x, " y=", y, " wx=", wx, " wy=", wy))
}
  


answer_d12p2 <- sum(abs(x), abs(y))
```

The answer to **Day 12 Part 2** is `r format(answer_d12p2, nsmall=0, big.mark=",")`

## Day 13: Shuttle Search
Your ferry can make it safely to a nearby port, but it won't get much further. When you call to book another ship, you discover that no ships embark from that port to your vacation island. You'll need to get from the port to the nearest airport.

Fortunately, a shuttle bus service is available to bring you from the sea port to the airport! Each bus has an ID number that also indicates how often the bus leaves for the airport....

[Full details....](https://adventofcode.com/2020/day/13)

### Day 13: Load the data
Note participants don't always get the same starting dataset so you should replace the day1.txt dataset with your own if you want to work out your answers.

```{r warning=FALSE, message=FALSE, day13_load}
#load the packages we need 
library(numbers)

#Remove previous day's data
rm(list = ls())

#Load the data that will be used for Part 1 & Part 2
#The data can be  downloaded from https://adventofcode.com/2020/day/13/
day13 <-readLines("data/day13.txt")


```

### Day 13: Part 1
What is the ID of the earliest bus you can take to the airport multiplied by the number of minutes you'll need to wait for that bus?

```{r day13_part1}
starttime<- as.integer(day13[1])
buses<-unlist(strsplit(day13[2],","))
buses<-as.integer(buses[buses!="x"])

nextbus <- function(bus, stime){
  if (numbers::mod(stime, bus)==0){
    #no wait
    return(stime)
  }else{
    return((floor(stime/bus)+1)*bus)
  }
}

#get next bus from start time
nxt <- sapply(buses, nextbus, stime=starttime)

answer_d13p1 <- ((min(nxt)-starttime)*buses[which.min(nxt)])
```

The answer to **Day 13 Part 1** is `r format(answer_d13p1, nsmall=0, big.mark=",")`

### Day 13: Part 2
The shuttle company is running a contest: one gold coin for anyone that can find the earliest timestamp such that the first bus ID departs at that time and each subsequent listed bus ID departs at that subsequent minute. (The first line in your input is no longer relevant.)...

...What is the earliest timestamp such that all of the listed bus IDs depart at offsets matching their positions in the list?

```{r day13_part2}
buses_temp<-unlist(strsplit(day13[2],","))
buses<-as.integer(buses_temp[buses_temp!="x"])

start_times <- vector("integer", length = length(buses))

#calculate start times
#subtracting the minute offset so that we are looking for busess that all have the same value
for (i in 1:length(buses)){
    start_times[i]<- -(match(buses[i], buses_temp)-1)
}

#setup initial values
bus29 <- c(0)
bus41 <- c(-19)
bus577 <- c(-29)
bus13 <- c(-42)
bus17 <- c(-43)
bus19 <- c(-48)
bus23 <- c(-52)
bus601 <- c(-60)
bus37 <- c(-97)

tmax <- 1000000000
tmax_interval <- 1000000000
answer_d13p2 <- as.double()

#brute force method works but is hugely inefficient for this combination of buses
#commented out as it takes so long to run

# while(length(answer_d13p2)==0){
#   bus29 <- seq(bus29[length(bus29)], tmax, by=29)
#   bus41 <- seq(bus41[length(bus41)], tmax, by=41)
#   bus577 <- seq(bus577[length(bus577)], tmax, by=577)
#   bus13 <- seq(bus13[length(bus13)], tmax, by=13)
#   bus17 <- seq(bus17[length(bus17)], tmax, by=17)
#   bus19 <- seq(bus19[length(bus19)], tmax, by=19)
#   bus23 <- seq(bus23[length(bus23)], tmax, by=23)
#   bus601 <- seq(bus601[length(bus601)], tmax, by=601)
#   bus37 <- seq(bus37[length(bus37)], tmax, by=37)
# 
#   answer_d13p2 <- Reduce(intersect, list(bus29, bus41, bus577, bus13, bus17, bus19, bus23, bus601, bus37))
#   tmax<- tmax + tmax_interval
#   message (paste("tmax = ", tmax))
# }

#answer_d13p2 <- start_times[1]
answer_d13p2 <-780601154795940

```

The answer to **Day 13 Part 2** is `r format(answer_d13p2, nsmall=0, big.mark=",")`

### Day 13: Part 2 - Better solution
I found AdroMine's solution while exploring more efficient ways of solving part 2.  It uses Chinese Remainder theorem to quickly solve the problem.  Much more efficient than my approach.

Check out [AdroMine's solution on Github](https://github.com/AdroMine/AdventOfCode2020/blob/main/Day13/solution.R)

```{r include=FALSE, EVAL = FALSE}

input <- readLines("data/day13.txt")

target <- as.numeric(input[1])
buses <- unlist(strsplit(input[2], ","))

# Part 1
bus_t <- as.numeric(buses[buses!="x"])
wait_time <- ((target %/% bus_t + 1) * bus_t) - target   # %/% is integer division in R

idx <- which.min(wait_time)
wait_time[idx] * bus_t[idx]


# for higher precision, wasted too much time because of 
# precision error
library(Rmpfr)

# Part 2
part2 <- function(buses){
    dept <- which(buses!="x") - 1
    buses <- as.double(buses[buses!="x"])
    
    .N <- function(.) mpfr(., precBits = 200)    
    N <- .N(prod(buses))
    Y <- N/buses
    rem <- -dept # can also use -dept %% buses
    inv <- rep(NA, length(buses))
    
    # inverse modulo
    for(i in 1:length(buses)){
        # n *k %% m equivalent to [(n %% m)*(k%%m)] %% m 
        m <- Y[i] %% buses[i]
        k <- 1:(buses[i]-1)
        remainder <- (m*k) %% buses[i]
        inv[i] <- which(remainder == 1)
    }
    sum(Y * rem * inv) %% N
}
part2(buses)

```

