---
title: "Advent of Code 2020"
author: "Barry Bullas"
output:
  html_document:
    toc: true
    toc_float: true
---
[Advent of Code](https://adventofcode.com/2020/about) is an Advent calendar of small programming puzzles for a variety of skill sets and skill levels that can be solved in any programming language you like. 

You don't need a computer science background to participate - just a little programming knowledge and some problem solving skills will get you pretty far. Nor do you need a fancy computer; every problem has a solution that completes in at most 15 seconds on ten-year-old hardware.

This project is my solution to those problems.  

You can view the code in my [advent2020 repository](https://github.com/barrychocolate/Advent2020/).

It is also published using Github Pages which can be viewed at [https://barrychocolate.github.io/Advent2020/solutions.html](https://barrychocolate.github.io/Advent2020/solutions.html)

### Setup
```{r initial_setup}
#Set this option so large numbers are not output in scientific notation
options(scipen=999)
```

## Day 1: Report Repair
Before you leave, the Elves in accounting just need you to fix your expense report (your puzzle input); apparently, something isn't quite adding up.  [Full details....](https://adventofcode.com/2020/day/1)

### Day 1: Load the data
Note participants don't always get the same starting dataset so you should replace the day1.txt dataset with your own if you want to work out your answers.

```{r warning=FALSE, message=FALSE, Day1_load}
#Load the data that will be used for Part 1 & Part 2
#The data can be  downloaded from https://adventofcode.com/2020/day/1/

day1 <- read.csv('data/day1.txt',header = FALSE)

#convert to vector
day1 <- unlist(day1)
```

### Day 1: Part 1
Specifically, they need you to find the two entries that sum to 2020 and then multiply those two numbers together.

```{r day1_part1}
#get all combinations of the numbers
day1_p1_combinations <- expand.grid(day1, day1)

#Calulate the sum of the two numbers so we can check if it equals 2020
day1_p1_combinations$agg <- day1_p1_combinations$Var1 + day1_p1_combinations$Var2

#Remove were the number is paired with itself
day1_p1_combinations <- day1_p1_combinations[day1_p1_combinations$Var1 != day1_p1_combinations$Var2, ]

#Deduplicate
day1_p1_combinations <- day1_p1_combinations[day1_p1_combinations$Var1 < day1_p1_combinations$Var2, ]

#Filter to find those with a sum of 2020
sum_to_2020_p1 <- day1_p1_combinations[day1_p1_combinations$agg==2020, ]

#Multiply the two numbers to get the answer
answer_d1p1 <- sum_to_2020_p1$Var1 * sum_to_2020_p1$Var2
```

The answer to **Day 1 Part 1** is `r format(answer_d1p1, nsmall=0, big.mark=",")`

### Day 1: Part 2

The Elves in accounting are thankful for your help; one of them even offers you a starfish coin they had left over from a past vacation. They offer you a second one if you can find three numbers in your expense report that meet the same criteria.

```{r day1_part2}
#get all combinations of the numbers
day1_p2_combinations <- expand.grid(day1, day1, day1)

#Calulate the sum of the two numbers so we can check if it equals 2020
day1_p2_combinations$agg <- day1_p2_combinations$Var1 + day1_p2_combinations$Var2 + day1_p2_combinations$Var3

#Remove were the number is paired with itself
day1_p2_combinations <- day1_p2_combinations[day1_p2_combinations$Var1 != day1_p2_combinations$Var2, ]

day1_p2_combinations <- day1_p2_combinations[day1_p2_combinations$Var2 != day1_p2_combinations$Var3, ]

#Deduplicate
day1_p2_combinations <- day1_p2_combinations[day1_p2_combinations$Var1 < day1_p2_combinations$Var2, ]

day1_p2_combinations <- day1_p2_combinations[day1_p2_combinations$Var2 < day1_p2_combinations$Var3, ]

#Filter to find those with a sum of 2020
sum_to_2020_p2 <- day1_p2_combinations[day1_p2_combinations$agg==2020, ]

#Multiply the two numbers to get the answer
#Only using the first record as the second record has the same numbers but alternated
answer_d1p2 <- sum_to_2020_p2[1,]$Var1 * sum_to_2020_p2[1,]$Var2 * sum_to_2020_p2[1,]$Var3
```

The answer to **Day 1 Part 2** is `r format(answer_d1p2, nsmall=0, big.mark=",")`

## Day 2: Password Philosophy
Your flight departs in a few days from the coastal airport; the easiest way down to the coast from here is via toboggan.

The shopkeeper at the North Pole Toboggan Rental Shop is having a bad day. "Something's wrong with our computers; we can't log in!" You ask if you can take a look.

Their password database seems to be a little corrupted: some of the passwords wouldn't have been allowed by the Official Toboggan Corporate Policy that was in effect when they were chosen.
[Full details....](https://adventofcode.com/2020/day/2)

### Day 2: Load the data
Note participants don't always get the same starting dataset so you should replace the day1.txt dataset with your own if you want to work out your answers.

```{r warning=FALSE, message=FALSE, day2_load}
#Remove previous days data
rm(list = ls())

#these librarys allow the use of the separate function and substr
library(tidyr, stringr)

#Load the data that will be used for Part 1 & Part 2
#The data can be  downloaded from https://adventofcode.com/2020/day/2/

day2 <- read.table('data/day2.txt', sep = " " , header = F, stringsAsFactors= F)
names(day2) <- c("char_count", "criteria", "pword")

#Remove : from criteria column
day2$criteria <- substr(day2$criteria, 1, 1)

#Split min and max and convert to int
day2 <- tidyr::separate(day2, char_count, c('mincount', 'maxcount'), sep = "-", convert = TRUE)


```

### Day 2: Part 1
To try to debug the problem, they have created a list (your puzzle input) of passwords (according to the corrupted database) and the corporate policy when that password was set.

For example, suppose you have the following list:
1-3 a: abcde
1-3 b: cdefg
2-9 c: ccccccccc

Each line gives the password policy and then the password. The password policy indicates the lowest and highest number of times a given letter must appear for the password to be valid. For example, 1-3 a means that the password must contain a at least 1 time and at most 3 times.

In the above example, 2 passwords are valid. The middle password, cdefg, is not; it contains no instances of b, but needs at least 1. The first and third passwords are valid: they contain one a or nine c, both within the limits of their respective policies.

How many passwords are valid according to their policies?

```{r day2_part1}
#Get a count of the number of times the criteria stirng appears in the password
day2$criteria_count <- stringr::str_count(day2$pword, day2$criteria)

#filter the dataset to includ only those where the count is 
day2_p1_valid <- day2 %>%
    dplyr::filter(criteria_count >= mincount & criteria_count <= maxcount)

answer_d2p1 <- nrow(day2_p1_valid)
```

The answer to **Day 2 Part 1** is `r format(answer_d2p1, nsmall=0, big.mark=",")`

### Day 2: Part 2
While it appears you validated the passwords correctly, they don't seem to be what the Official Toboggan Corporate Authentication System is expecting.

The shopkeeper suddenly realizes that he just accidentally explained the password policy rules from his old job at the sled rental place down the street! The Official Toboggan Corporate Policy actually works a little differently.

Each policy actually describes two positions in the password, where 1 means the first character, 2 means the second character, and so on. (Be careful; Toboggan Corporate Policies have no concept of "index zero"!) Exactly one of these positions must contain the given letter. Other occurrences of the letter are irrelevant for the purposes of policy enforcement.

How many passwords are valid according to the new interpretation of the policies?

```{r day2_part2}
#extract the characters at the positions
day2$charpos1 <-substr(day2$pword, start = day2$mincount, stop = day2$mincount)
day2$charpos2 <-substr(day2$pword, start = day2$maxcount, stop = day2$maxcount)

#Filter toidentify the valid passwords
#i.e the character is in one of the two positions but not both
day2_p2_valid <- day2 %>%
    dplyr::filter( (charpos1 == criteria | charpos2 == criteria) & charpos1 != charpos2) 

#Get a count of the number of valid passowrds
answer_d2p2 <- nrow(day2_p2_valid)
```

The answer to **Day 2 Part 2** is `r format(answer_d2p2, nsmall=0, big.mark=",")`

## Day 3: Toboggan Trajectory
With the toboggan login problems resolved, you set off toward the airport. While travel by toboggan might be easy, it's certainly not safe: there's very minimal steering and the area is covered in trees. You'll need to see which angles will take you near the fewest trees.

Due to the local geology, trees in this area only grow on exact integer coordinates in a grid. You make a map (your puzzle input) of the open squares (.) and trees (#) you can see.

[Full details....](https://adventofcode.com/2020/day/3)

### Day 3: Load the data
Note participants don't always get the same starting dataset so you should replace the day1.txt dataset with your own if you want to work out your answers.

```{r warning=FALSE, message=FALSE, day3_load}
#Remove previous day's data
rm(list = ls())

#dplyr for filter, numebrs for mod function
library(dplyr, numbers)

#Load the data that will be used for Part 1 & Part 2
#The data can be  downloaded from https://adventofcode.com/2020/day/3/

day3 <- read.csv('data/day3.txt',header = FALSE, stringsAsFactors = FALSE, col.names = c('map'))
```

### Day 3: Part 1
Starting at the top-left corner of your map and following a slope of right 3 and down 1, how many trees would you encounter?

```{r day3_part1}
#create a function to count the number of trees
#I have used a function as Advent of Code is likely going to ask me to do something similar in Part 2

howmanytrees <- function(data, starting_position, rmove, dmove, maprowlength){
#create a temporary copy of the data for use within the function
temp_data<-data

#Remove rows we won't land on
temp_data <- temp_data %>%
    dplyr::filter (numbers::mod(1:nrow(temp_data)-1,dmove)==0)

#Calculate the position on the map where the toboggan will land
temp_data$position <- numbers::mod((starting_position +((1:nrow(temp_data)-1)*rmove)), maprowlength)

#update any position zero to the maprowlength
temp_data$position[temp_data$position ==0] <- maprowlength

#Use the position to calculate what is at that position and store it in the target column
temp_data$target <- substr(temp_data$map, start = temp_data$position, stop = temp_data$position)

#Get a count of all targets that were trees
#Filter rows to get only those that would be targeted based on the downward movement
return(as.numeric(nrow(temp_data %>% dplyr::filter (target == '#'))))
}

#Call the function with the day 3 data, starting at position 1, with a movement of 3 and a map row length of 31
answer_d3p1<-howmanytrees(day3, 1, 3, 1, 31)

```

The answer to **Day 3 Part 1** is `r format(answer_d3p1, nsmall=0, big.mark=",")`

### Day 3: Part 2
Time to check the rest of the slopes - you need to minimize the probability of a sudden arboreal stop, after all.

Determine the number of trees you would encounter if, for each of the following slopes, you start at the top-left corner and traverse the map all the way to the bottom:

Right 1, down 1.
Right 3, down 1. (This is the slope you already checked.)
Right 5, down 1.
Right 7, down 1.
Right 1, down 2.
In the above example, these slopes would find 2, 7, 3, 4, and 2 tree(s) respectively; multiplied together, these produce the answer 336.

```{r day3_part2}
#Call the function for the Part 2 requirements
r1d1 <- howmanytrees(day3, 1, 1, 1, 31)
r3d1 <- howmanytrees(day3, 1, 3, 1, 31)
r5d1 <- howmanytrees(day3, 1, 5, 1, 31)
r7d1 <- howmanytrees(day3, 1, 7, 1, 31)
r1d2 <- howmanytrees(day3, 1, 1, 2, 31) 

#multiply the results together to get the answer to part 2
answer_d3p2 <- r1d1 * r3d1 * r5d1 * r7d1 * r1d2
```

The answer to **Day 3 Part 2** is `r format(answer_d3p2, nsmall=0, big.mark=",")`

## Day 4: Passport Processing
You arrive at the airport only to realize that you grabbed your North Pole Credentials instead of your passport. While these documents are extremely similar, North Pole Credentials aren't issued by a country and therefore aren't actually valid documentation for travel in most of the world.

It seems like you're not the only one having problems, though; a very long line has formed for the automatic passport scanners, and the delay could upset your travel itinerary.

Due to some questionable network security, you realize you might be able to solve both of these problems at the same time.

The automatic passport scanners are slow because they're having trouble detecting which passports have all required fields. 

[Full details....](https://adventofcode.com/2020/day/1)

### Day 4: Load the data
Note participants don't always get the same starting dataset so you should replace the day1.txt dataset with your own if you want to work out your answers.

```{r warning=FALSE, message=FALSE, day4_load}
#load the packages we need
library(stringr)
library(tidyr)
library(dplyr)

#Remove previous day's data
rm(list = ls())

#Load the data that will be used for Part 1 & Part 2
#The data can be  downloaded from https://adventofcode.com/2020/day/4/
day4 <-  readLines('data/day4.txt')
nvec <- length(day4)

#Identify empty lines between data as a record terminator
#remove new lines in the key value pairs within records
#So all the data for each record is in one string
breaks <- which(! nzchar(day4))
nbreaks <- length(breaks)
if (breaks[nbreaks] < nvec) {
  breaks <- c(breaks, nvec + 1L)
  nbreaks <- nbreaks + 1L
}
if (nbreaks > 0L) {
  day4 <- mapply(function(a,b) paste(day4[a:b], collapse = " "),
                   c(1L, 1L + breaks[-nbreaks]),
                   breaks - 1L)
}
day4<- as.data.frame(day4)
names(day4)<- c('vpairs')

#Seperate the key value pairs out into columns
KVsep <- fixed(" ")  #key-value separator
Vsep <- fixed(":")     #value separator

day4 <-  day4 %>%
  mutate(KVpairs = str_split(vpairs, KVsep)) %>%
  unnest(KVpairs) %>%
  separate(KVpairs, into = c("key", "value"), Vsep) %>%
  spread(key, value) %>%
  select(-vpairs)

#remove data that is no longer needed
rm(breaks, KVsep, nbreaks, nvec, Vsep)
```

### Day 4: Part 1
Passport data is validated in batch files (your puzzle input). Each passport is represented as a sequence of key:value pairs separated by spaces or newlines. Passports are separated by blank lines.

According to the above rules, your improved system would report 2 valid passports.

Count the number of valid passports - those that have all required fields. Treat cid as optional. In your batch file, how many passports are valid?

```{r day4_part1}
#replace missing CID values with NPC (North Pole Credentials)
#This will make it easier to identify records that have NAs in any other column
day4$cid[is.na(day4$cid)]<-'NPC'

day4_no_na <-day4[!rowSums(is.na(day4[names(day4)])), ]

#Count rows that don't contain NA
answer_d4p1<- nrow(day4_no_na)
```

The answer to **Day 4 Part 1** is `r format(answer_d4p1, nsmall=0, big.mark=",")`

### Day 4: Part 2
The line is moving more quickly now, but you overhear airport security talking about how passports with invalid data are getting through. Better add some data validation, quick!

You can continue to ignore the cid field, but each other field has strict rules about what values are valid for automatic validation:

* byr (Birth Year) - four digits; at least 1920 and at most 2002.
* iyr (Issue Year) - four digits; at least 2010 and at most 2020.
* eyr (Expiration Year) - four digits; at least 2020 and at most 2030.
* hgt (Height) - a number followed by either cm or in:
* If cm, the number must be at least 150 and at most 193.
* If in, the number must be at least 59 and at most 76.
* hcl (Hair Color) - a # followed by exactly six characters 0-9 or a-f.
* ecl (Eye Color) - exactly one of: amb blu brn gry grn hzl oth.
* pid (Passport ID) - a nine-digit number, including leading zeroes.
* cid (Country ID) - ignored, missing or not.

```{r day4_part2}
#Filter to remove invalid birth years that are not 4 digits between 1920 and 2002
day4_no_na <- day4_no_na %>%
  filter(grepl('^[[:digit:]]{4}$', byr) & between(as.integer(byr), 1920, 2002))

#Filter to remove invalid issue years that are not 4 digits between 2010 and 2020
day4_no_na <- day4_no_na %>%
  filter(grepl('^[[:digit:]]{4}$', iyr) & between(as.integer(iyr), 2010, 2020))

#Filter to remove expiration years that are not 4 digits between 2020 and 2030
day4_no_na <- day4_no_na %>%
  filter(grepl('^[[:digit:]]{4}$', eyr) & between(as.integer(eyr), 2020, 2030))

#Filter to remove invalid height 
#extract height to a column as an integer
day4_no_na$height <-  as.integer(str_sub(day4_no_na$hgt, 1, -3))

#extract the height type (in or cm) to a column
day4_no_na$heighttype <-  str_sub(day4_no_na$hgt, -2, -1)

day4_no_na <- day4_no_na %>%
  filter((heighttype == 'cm' & between(height, 150, 193)) | (heighttype == 'in' & between(height, 59, 76)))

#Filter to remove invalid hair colour -  a # followed by exactly six characters 0-9 or a-f.
day4_no_na <- day4_no_na %>%
  filter(grepl('^#[0-9a-f]{6}$', hcl))

#Filter to remove invalid eye colour - exactly one of: amb blu brn gry grn hzl oth.
valid_eyes <- c('amb', 'blu', 'brn', 'gry', 'grn', 'hzl', 'oth')

day4_no_na <- day4_no_na %>%
  filter(ecl %in% valid_eyes)

#Filter to remove invalid passport id - a nine-digit number, including leading zeroes
day4_no_na <- day4_no_na %>%
  filter(grepl('^[0-9]{9}$', pid))


answer_d4p2 <- nrow(day4_no_na)
```

The answer to **Day 4 Part 2** is `r format(answer_d4p2, nsmall=0, big.mark=",")`

## Day 5: Binary Boarding
You board your plane only to discover a new problem: you dropped your boarding pass! You aren't sure which seat is yours, and all of the flight attendants are busy with the flood of people that suddenly made it through passport control.

You write a quick program to use your phone's camera to scan all of the nearby boarding passes (your puzzle input); perhaps you can find your seat through process of elimination.

Instead of zones or groups, this airline uses binary space partitioning to seat people. A seat might be specified like FBFBBFFRLR, where F means "front", B means "back", L means "left", and R means "right".

The first 7 characters will either be F or B; these specify exactly one of the 128 rows on the plane (numbered 0 through 127). Each letter tells you which half of a region the given seat is in. Start with the whole list of rows; the first letter indicates whether the seat is in the front (0 through 63) or the back (64 through 127). The next letter indicates which half of that region the seat is in, and so on until you're left with exactly one row.

[Full details....](https://adventofcode.com/2020/day/1)

### Day 5: Load the data
Note participants don't always get the same starting dataset so you should replace the day1.txt dataset with your own if you want to work out your answers.

```{r warning=FALSE, message=FALSE, day5_load}
#load the packages we need
library(dplyr)

#Remove previous day's data
rm(list = ls())

#Load the data that will be used for Part 1 & Part 2
#The data can be  downloaded from https://adventofcode.com/2020/day/5/
day5 <-read.table('data/day5.txt',sep="",header=FALSE, col.names = c('bsp'), stringsAsFactors = FALSE)

```

### Day 5: Part 1
Every seat also has a unique seat ID: multiply the row by 8, then add the column. In this example, the seat has ID 44 * 8 + 5 = 357.

Here are some other boarding passes:

* BFFFBBFRRR: row 70, column 7, seat ID 567.
* FFFBBBFRRR: row 14, column 7, seat ID 119.
* BBFFBBFRLL: row 102, column 4, seat ID 820.

As a sanity check, look through your list of boarding passes. What is the highest seat ID on a boarding pass?

```{r day5_part1}
#create a function to calculate the SeatID
calculateSeatId <- function(seatRow, seatCol){
  return (as.integer((seatRow * 8)+seatCol))
}
  
  
findSeat <- function(binarySpacePartitioning){
  #setup a vector for the seat row and column
  seatRow<-c(0:127)
  seatCol<-c(0:7)
  
  #extract the row and col positions from the bsp into vectors
  rowPos <-strsplit(substr(binarySpacePartitioning, start = 1, stop = 7),"")[[1]]
  colPos <-strsplit(substr(binarySpacePartitioning, start = 8, stop = 10),"")[[1]]
  
  seatRow<-c(0:127)
  seatCol<-c(0:7)
  
  #loop through each row position
  for (bsprow in rowPos) {
    if (bsprow == 'F') {
      #take first part of vector
      seatRow<- seatRow[1:(length(seatRow)/2)]
    } else {
      #take second part of vector
      seatRow<- seatRow[((length(seatRow)/2)+1):(length(seatRow))]
    }
  }
  
  #loop through each column position
  for (bspcol in colPos) {
    if (bspcol == 'L') {
      #take first part of vector
      seatCol<- seatCol[1:(length(seatCol)/2)]
    } else {
      #take second part of vector
      seatCol<- seatCol[((length(seatCol)/2)+1):(length(seatCol))]
    }
  }
  return (calculateSeatId(seatRow, seatCol))
}

#Apply the function to all binary Space Partitioning identifiers
day5$seatID <-  sapply(day5$bsp, findSeat)

#order by seat ID
day5 <- day5[order(day5$seatID),]

#Get the maximum seatID
answer_d5p1<- max (day5$seatID, na.rm = TRUE)
```

The answer to **Day 5 Part 1** is `r format(answer_d5p1, nsmall=0, big.mark=",")`

### Day 5: Part 2
Ding! The "fasten seat belt" signs have turned on. Time to find your seat.

It's a completely full flight, so your seat should be the only missing boarding pass in your list. However, there's a catch: some of the seats at the very front and back of the plane don't exist on this aircraft, so they'll be missing from your list as well.

Your seat wasn't at the very front or back, though; the seats with IDs +1 and -1 from yours will be in your list.

```{r day5_part2}
#create a vector or all rows skipping the first and last row
allRow<-c(1:126)

#create a vector of all columns
allCol<-c(0:7)

#great all possible seat combinations using the two vectors
allSeats<-expand.grid(seatRow = allRow, seatCol =allCol)

#calculate the seat id of all possible seats
allSeats$seatID <- calculateSeatId(allSeats$seatRow, allSeats$seatCol)

#order by seat id
allSeats <- allSeats[order(allSeats$seatID),]

#Find the missing values between all possible seats and actual seats
missingSeats <- as.data.frame(setdiff(allSeats$seatID,day5$seatID))
names(missingSeats) <- 'seatId'

#calculate the id of the seat +1 and -1
missingSeats$oneup <- missingSeats$seatId + 1
missingSeats$onedown <- missingSeats$seatId -1

#remove any seats where seatid + 1 or -1 exists (it shouldn't if it is your seat)
missingSeats  <- missingSeats %>%
  filter(!(missingSeats$oneup %in% missingSeats$seatId) & !(missingSeats$onedown %in% missingSeats$seatId))

answer_d5p2 <- missingSeats$seatId
```

The answer to **Day 5 Part 2** is `r format(answer_d5p2, nsmall=0, big.mark=",")`

## Day 6: Custom Customs
As your flight approaches the regional airport where you'll switch to a much larger plane, customs declaration forms are distributed to the passengers.

The form asks a series of 26 yes-or-no questions marked a through z. All you need to do is identify the questions for which anyone in your group answers "yes". Since your group is just you, this doesn't take very long.

However, the person sitting next to you seems to be experiencing a language barrier and asks if you can help. For each of the people in their group, you write down the questions for which they answer "yes", one per line.

[Full details....](https://adventofcode.com/2020/day/6)

### Day 6: Load the data
Note participants don't always get the same starting dataset so you should replace the day1.txt dataset with your own if you want to work out your answers.

```{r warning=FALSE, message=FALSE, day6_load}
#load the packages we need 
library(purrr) #for map2 and map_dbl

#Remove previous day's data
rm(list = ls())

#Load the data that will be used for Part 1 & Part 2
#The data can be  downloaded from https://adventofcode.com/2020/day/6/
day6 <- readLines("data/day6.txt")

# Find where groups start and end
new_lines <- which(day6 == '')
group_start <- c(1, new_lines + 1)
group_end <- c(new_lines - 1, length(day6))

# Create a list of customs cards for traveller groups
day6 <- map2(group_start, group_end, ~day6[.x : .y])

#remove data no longer required
rm(group_start, group_end, new_lines)

```

### Day 6: Part 1
For each group, count the number of questions to which anyone answered "yes". What is the sum of those counts?

```{r day6_part1}
#1 paste combines all items in each list to one string
#2 strsplit splits the combined string into a list of characters
#3 unique provides a list of all characters that appear in the string
#4 Length gets the length of the unique characters i.e the number of unique questions answered
#Map_dbl allows us to apply this nested function to each element in the list of grouped customs cards
answer_d6p1<- sum(map_dbl(day6, ~length(unique(strsplit(paste(., collapse = ""), "")[[1]]))))
```

The answer to **Day 6 Part 1** is `r format(answer_d6p1, nsmall=0, big.mark=",")`

### Day 6: Part 2
s you finish the last group's customs declaration, you notice that you misread one word in the instructions:

You don't need to identify the questions to which anyone answered "yes"; you need to identify the questions to which everyone answered "yes"!

For each group, count the number of questions to which everyone answered "yes". What is the sum of those counts?

```{r day6_part2}
#strsplit splits the elements within a character string or vector into a list
#reduce allows us to apply the intersect function to that list to retain only elements present in each list i.e the questions everyone answered
#length counts eachthe number of questions everyone answered
#sum aggregates that for all travellers
answer_d6p2 <- sum(map_dbl(day6, function(x) length(reduce(strsplit(x, ""), intersect))))
```

The answer to **Day 6 Part 2** is `r format(answer_d6p2, nsmall=0, big.mark=",")`