---
title: "Advent of Code 2020"
author: "Barry Bullas"
output:
  html_document:
    toc: true
    toc_float: true
---

[Advent of Code](https://adventofcode.com/2020/about) is an Advent calendar of small programming puzzles for a variety of skill sets and skill levels that can be solved in any programming language you like. 

This project is my solution to those problems.

## Day 1: Report Repair
Before you leave, the Elves in accounting just need you to fix your expense report (your puzzle input); apparently, something isn't quite adding up.  [Full details....](https://adventofcode.com/2020/day/1)

### Load the data
Note participants don't always get the same starting dataset so you should replace the day1.txt dataset with your own if you want to work out your answers.

```{r}
#Load the data that will be used for Part 1 & Part 2
#The data can be  downloaded from https://adventofcode.com/2020/day/1/

day1 <- read.csv('data/day1.txt',header = FALSE)

#convert to vector
day1 <- unlist(day1)
```

### Part 1
Specifically, they need you to find the two entries that sum to 2020 and then multiply those two numbers together.

```{r}
#get all combinations of the numbers
day1_p1_combinations <- expand.grid(day1, day1)

#Calulate the sum of the two numbers so we can check if it equals 2020
day1_p1_combinations$agg <- day1_p1_combinations$Var1 + day1_p1_combinations$Var2

#Filter to find those with a sum of 2020
sum_to_2020_p1 <- day1_p1_combinations[day1_p1_combinations$agg==2020, ]

#Multiply the two numbers to get the answer
#Only using the first record as the second record has the same numbers but alternated
answer1 <- sum_to_2020_p1[1,]$Var1 * sum_to_2020_p1[1,]$Var2

```

The answer to **Part One** is `r answer1`

### Part 2

The Elves in accounting are thankful for your help; one of them even offers you a starfish coin they had left over from a past vacation. They offer you a second one if you can find three numbers in your expense report that meet the same criteria.

```{r}
#get all combinations of the numbers
day1_p2_combinations <- expand.grid(day1, day1, day1)

#Calulate the sum of the two numbers so we can check if it equals 2020
day1_p2_combinations$agg <- day1_p2_combinations$Var1 + day1_p2_combinations$Var2 + day1_p2_combinations$Var3

#Filter to find those with a sum of 2020
sum_to_2020_p2 <- day1_p2_combinations[day1_p2_combinations$agg==2020, ]

#Multiply the two numbers to get the answer
#Only using the first record as the second record has the same numbers but alternated
answer2 <- sum_to_2020_p2[1,]$Var1 * sum_to_2020_p2[1,]$Var2 * sum_to_2020_p2[1,]$Var3
```

The answer to **Part Two** is `r answer2`
