---
title: "Advent of Code 2020"
author: "Barry Bullas"
output:
  html_document:
    toc: true
    toc_float: true
---

[Advent of Code](https://adventofcode.com/2020/about) is an Advent calendar of small programming puzzles for a variety of skill sets and skill levels that can be solved in any programming language you like. 

This project is my solution to those problems.  

You can view the code in my [advent2020 repository](https://github.com/barrychocolate/Advent2020/).

It is also published using Github Pages which can be viewed at [https://barrychocolate.github.io/Advent2020/solutions.html](https://barrychocolate.github.io/Advent2020/solutions.html)

## Day 1: Report Repair
Before you leave, the Elves in accounting just need you to fix your expense report (your puzzle input); apparently, something isn't quite adding up.  [Full details....](https://adventofcode.com/2020/day/1)

### Load the data
Note participants don't always get the same starting dataset so you should replace the day1.txt dataset with your own if you want to work out your answers.

```{r Day1_load}
#Load the data that will be used for Part 1 & Part 2
#The data can be  downloaded from https://adventofcode.com/2020/day/1/

day1 <- read.csv('data/day1.txt',header = FALSE)

#convert to vector
day1 <- unlist(day1)
```

### Part 1
Specifically, they need you to find the two entries that sum to 2020 and then multiply those two numbers together.

```{r day1_part1}
#get all combinations of the numbers
day1_p1_combinations <- expand.grid(day1, day1)

#Calulate the sum of the two numbers so we can check if it equals 2020
day1_p1_combinations$agg <- day1_p1_combinations$Var1 + day1_p1_combinations$Var2

#Remove were the number is paired with itself
day1_p1_combinations <- day1_p1_combinations[day1_p1_combinations$Var1 != day1_p1_combinations$Var2, ]

#Deduplicate
day1_p1_combinations <- day1_p1_combinations[day1_p1_combinations$Var1 < day1_p1_combinations$Var2, ]

#Filter to find those with a sum of 2020
sum_to_2020_p1 <- day1_p1_combinations[day1_p1_combinations$agg==2020, ]

#Multiply the two numbers to get the answer
answer_d1p1 <- sum_to_2020_p1$Var1 * sum_to_2020_p1$Var2

#remove datasets no longer needed
rm(day1_p1_combinations, sum_to_2020_p1)
```

The answer to **Day 1 Part 1** is `r answer_d1p1`

### Part 2

The Elves in accounting are thankful for your help; one of them even offers you a starfish coin they had left over from a past vacation. They offer you a second one if you can find three numbers in your expense report that meet the same criteria.

```{r day1_part2}
#get all combinations of the numbers
day1_p2_combinations <- expand.grid(day1, day1, day1)

#Calulate the sum of the two numbers so we can check if it equals 2020
day1_p2_combinations$agg <- day1_p2_combinations$Var1 + day1_p2_combinations$Var2 + day1_p2_combinations$Var3

#Remove were the number is paired with itself
day1_p2_combinations <- day1_p2_combinations[day1_p2_combinations$Var1 != day1_p2_combinations$Var2, ]

day1_p2_combinations <- day1_p2_combinations[day1_p2_combinations$Var2 != day1_p2_combinations$Var3, ]

#Deduplicate
day1_p2_combinations <- day1_p2_combinations[day1_p2_combinations$Var1 < day1_p2_combinations$Var2, ]

day1_p2_combinations <- day1_p2_combinations[day1_p2_combinations$Var2 < day1_p2_combinations$Var3, ]

#Filter to find those with a sum of 2020
sum_to_2020_p2 <- day1_p2_combinations[day1_p2_combinations$agg==2020, ]

#Multiply the two numbers to get the answer
#Only using the first record as the second record has the same numbers but alternated
answer_d1p2 <- sum_to_2020_p2[1,]$Var1 * sum_to_2020_p2[1,]$Var2 * sum_to_2020_p2[1,]$Var3

#remove datasets no longer needed
rm(day1_p2_combinations, sum_to_2020_p2, day1)
```

The answer to **Day 1 Part 2** is `r answer_d1p2`
